### 2019-02-11 benchmarking/paper/draft0.Rmd
merged in benchmarking into v3b. changed 3.3v power pin so that design is compatible with teensy 3.1/2 and 3.5/6.

---
title: "Microstim: open source hardware for precise electrical stimulation"
author: ""
date: "February 11, 2019"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Microstim is built like so
  - Two independently-controllable output channels, each with both current and voltage output modes
  - Total cost: $131 USD.
  - Output range in current mode: -833uA to +833uA.
  - Pulse width in current mode: down to 50 us.
  - Output range in voltage mode: -12V to 12V.
  - Pulse width in current mode: down to 10 us.
  - In current mode, +-11V compliance voltage (+/-833uA output possible for resistances up to 13k, +-110uA possible for resistances up to 100k)
  - Channels are isolated 
  - Onboard ADC measurement of actual output current or voltage
  - runs on USB power
  - uses a Teensy-based microcontroller
  - FIGURE 1 (picture)

## Microstim is precise

  - amplitude accuracy
    - voltage and current output
  - pulse timing accuracy
  - channels are isolated
  - FIGURE 2 (benchmarks)

## Microstim is digital and highly configurable

## Microstim is better than everything else
  - TABLE 1 comparing stimulators?
    - STG3000 / 4000 from multichannel systems
    - WPI A310 Accupulser?
    - pulsepal

## Microstim limitations

## Potential applications

  - in vivo stimulation

## Supplemental materials
  - BOM
  - gerbers
  - schematic
  - link to bitbucket

### 2019-02-12 benchmarking/paper/draft0.Rmd
updates to benchmarking parameter sets and analysis code

@@
 output: html_document
+editor_options: 
+  chunk_output_type: console
 ---
@@
 ```{r setup, include=FALSE}
-knitr::opts_chunk$set(echo = TRUE)
+knitr::opts_chunk$set(echo = FALSE)
+library(parallel)
 ```
+
+```{r loadData, cache=TRUE}
+
+projectDirectory = '/home/x/Desktop/microstim/benchmarking/'
+setwd(projectDirectory)
+
+BOM=openxlsx::read.xlsx("../microstim_BOM.xlsx")
+sum(BOM$Unit.cost*BOM$Quantity, na.rm=TRUE)
+
+paramMat = read.csv(paste(projectDirectory,"paramMat.csv", sep=''))
+colnames(paramMat) = c('mode0','mode1','period','duration','ch0amp0','ch1amp0','dur0', 'ch0amp1','ch1amp1','dur1')
+
+vpau = 20/2^12 # volts per adc unit
+fs = 2e6       # sampling rate of measurements
+
+dataFiles = list.files(path=paste(projectDirectory,'data',sep=''), pattern="[.]bin", full.names = TRUE)
+
+# readTestFile = function(i){
+#   sapply(dataFiles[i], function(f) {
+#     matrix(readBin(f, what='integer', n=1e7, size=2, endian = 'big'),ncol=2,byrow=TRUE)
+#   }, simplify='array')
+# }
+
+getPulseParams = function(f, thresh=100, fs, vpau){
+
+  dat = matrix(readBin(f, what='integer', n=1e7, size=2, endian = 'big'),ncol=2,byrow=TRUE)
+
+  apply(dat, 2, function(x) {  
+    isPulse = 1.0*(x>thresh | x < -thresh)
+    isPulse = mmand::erode(mmand::dilate(isPulse, rep(1,21)),rep(1,21))
+    pulseIndices = which(isPulse>0)
+
+    pulseStartIndices = pulseIndices[c(1,which(diff(pulseIndices) > 3)+1)]
+    pulseEndIndices = pulseIndices[c(which(diff(pulseIndices) > 3), length(pulseIndices))]
+
+    isStageHigh = 1.0*(x > thresh)
+    isStageLow = 1.0*(x < -thresh)
+
+    plot(x[1:3000], type='l', ylim=c(190,220))
+    points(pulseIndices, x[pulseIndices], cex=0.3, col='red')
+    points(pulseStartIndices, x[pulseStartIndices], cex=1, col='blue')
+    points(pulseEndIndices, x[pulseEndIndices], cex=1, col='orange')
+
+    IPI = diff(pulseStartIndices/fs)
+    IPI.summary = summary(IPI)
+    IPI.sd = sd(IPI)
+
+    PW = (pulseEndIndices-pulseStartIndices)/fs
+    PW.summary = summary(PW)
+    PW.sd = sd(PW)
+
+    sh.median = median(x[which(isStageHigh>0)])*vpau
+    sh.sd = sd(x[which(mmand::erode(isStageHigh, rep(1,13))>0)])*vpau
+    sl.median = median(x[which(isStageLow>0)])*vpau
+    sl.sd = sd(x[which(mmand::erode(isStageLow, rep(1,13))>0)])*vpau
+
+    list(IPI.summary=IPI.summary, IPI.sd=IPI.sd, PW.summary=PW.summary, 
+             PW.sd=PW.sd, sh.median=sh.median, sh.sd=sh.sd, 
+             sl.median=sl.median, sl.sd=sl.sd)
+  })
+}
+
+j = which(paramMat$mode0 ==0 & paramMat$mode1==2)
+
+cl = makeCluster(detectCores()-1)
+res = parLapply(cl, dataFiles[j], getPulseParams, fs=fs, vpau=vpau)
+res = unlist(res, recursive = FALSE)
+stopCluster(cl)
+```
+
+To benchmark, we varied mode, pulse frequency (from `r round(1e6/max(paramMat$period),2)`Hz to `r round(1e6/min(paramMat$period),2)`kHz), pulse duration (from `r min(paramMat$dur0)` to `r max(paramMat$dur0)` us), and amplitude.
+We simultaneously recorded from both output channels on a National Instruments PCI-6110 card (12-bit ADC, sampling each channel at 2MHz).
 
@@
   - Two independently-controllable output channels, each with both current and voltage output modes
-  - Total cost: $131 USD.
+  - Total cost: $`r ceiling(sum(BOM$Unit.cost*BOM$Quantity, na.rm=TRUE))` USD.
   - Output range in current mode: -833uA to +833uA.
@@
   - FIGURE 2 (benchmarks)
+
+```{r benchmarkFigure, fig.height=8,fig.width=8}
+
+par(mar=c(4,5,2,2), mfrow=c(4,3))
+plot(sapply(res, function(x) x$IPI.summary[3])*1e6, ylab='Median interpulse interval\n(microsec)')
+abline(h=2*unique(paramMat$period), col='red')
+plot(sapply(res, function(x) x$IPI.sd)*1e6, ylab='Interpulse interval stdev\n(microsec)')
+abline(h=0.5)
+plot(sapply(res, function(x) diff(x$IPI.summary[c(1,6)]))*1e6, ylab='Interpulse interval range\n(microsec)')
+abline(h=0.5)
+
+plot(sapply(res, function(x) x$PW.summary[3])*1e6, ylab='Median pulse width\n(microsec)')
+abline(h=2*unique(paramMat$dur0), col='red')
+plot(sapply(res, function(x) x$PW.sd)*1e6, ylab='Pulse width stdev\n(microsec)')
+abline(h=0.5)
+plot(sapply(res, function(x) diff(x$PW.summary[c(1,6)]))*1e6, ylab='Pulse width range\n(microsec)')
+abline(h=0.5)
+
+plot(sapply(res, function(x) x$sh.median), ylab='Median pos. pulse amplitude (V)')
+plot(sapply(res, function(x) x$sh.sd), ylab='Pulse amplitude sd (V)')
+abline(h=0.5)
+plot(1,1)
+
+plot(sapply(res, function(x) x$sl.median), ylab='Median neg. pulse amplitude (V)')
+plot(sapply(res, function(x) x$sl.sd), ylab='Pulse amplitude sd (V)')
+abline(h=0.5)
+plot(1,1)
+
+```
 

### 2019-02-12 benchmarking/makeTestParameters.R

@@ -1,21 +1,24 @@
+# Generate test parameters for benchmarking microstim 
+# (c) Nathan Cermak 2019
+#
+#
+
+projectDirectory = '/home/x/Desktop/microstim/'
 
 duration = 1e6
-
 paramMat = NULL
+mode0 = 0
+mode1 = 2
+amp = c(1000,100,100,100)[c(mode0,mode1)+1]
+ampScalings = c(1,5)
+pulseDurations = c(20,50,100,200,500,1000,2000)
+periods = round(exp(seq(log(500), log(5e5), length.out=10)))
 
-for (mode1 in 0:3){
-  for(mode2 in 0:3){
-    for (period in round(exp(seq(log(500), log(5e5), length.out=20)))){
-      amp = c(1000,100,100,100)[c(mode1,mode2)+1]
-      for (pulseDuration in c(20,30,40,50,75,100,200)){
-        params = c(mode1, mode2, period, duration, amp, pulseDuration, -amp, pulseDuration)
-        paramMat = rbind(paramMat, params)        
-      }
-      amp = amp*5
-      for (pulseDuration in c(20,30,40,50,75,100,200)){
-        params = c(mode1, mode2, period, duration, amp, pulseDuration, -amp, pulseDuration)
-        paramMat = rbind(paramMat, params)        
-      }
+for (period in periods){
+  for (pulseDuration in pulseDurations[pulseDurations<period/2]){
+    for (ampScaling in ampScalings){
+      params = c(mode0, mode1, period, duration, amp*ampScaling, pulseDuration, -amp*ampScaling, pulseDuration)
+      paramMat = rbind(paramMat, params)        
     }
   }
 }
@@ -25,24 +28,26 @@ par2string = function(p){
 }
 
 stringVector = apply(paramMat,1,par2string)
-cat(stringVector, file="/home/x/Desktop/paramStringFile.txt")
-write.csv(paramMat, "/home/x/Desktop/paramMat.csv")
+cat(stringVector, file=paste(projectDirectory,"benchmarking/paramStringFile.txt", sep=''))
+write.csv(paramMat, paste(projectDirectory,"benchmarking/paramMat.csv", sep=''),row.names = FALSE)
 
-con <- serial::serialConnection(name = "testcon",port = "ttyACM0",
-                        mode = "256000,n,8,1",
-                        newline = 0,
-                        translation = "auto")
-open(con)
-serial::read.serialConnection(con)
 
[…]

### 2019-02-25 benchmarking/paper/draft0.Rmd
boatload of updates via jons suggestions. de-isolated grounds via large resistors+capacitors (to prevent static buildup on animal or issues with large discharges). increased resistor values in voltage output non-inverting amplifier.

@@
 
-    plot(x[1:3000], type='l', ylim=c(190,220))
-    points(pulseIndices, x[pulseIndices], cex=0.3, col='red')
-    points(pulseStartIndices, x[pulseStartIndices], cex=1, col='blue')
-    points(pulseEndIndices, x[pulseEndIndices], cex=1, col='orange')
+    #plot(x[1:3000], type='l', ylim=c(190,220))
+    #points(pulseIndices, x[pulseIndices], cex=0.3, col='red')
+    #points(pulseStartIndices, x[pulseStartIndices], cex=1, col='blue')
+    #points(pulseEndIndices, x[pulseEndIndices], cex=1, col='orange')
 
@@
 
+    pulseMat = sapply(pulseStartIndices, function(j) x[j+(-100:1000)])
     list(IPI.summary=IPI.summary, IPI.sd=IPI.sd, PW.summary=PW.summary, 
              PW.sd=PW.sd, sh.median=sh.median, sh.sd=sh.sd, 
-             sl.median=sl.median, sl.sd=sl.sd)
+             sl.median=sl.median, sl.sd=sl.sd, pulseMat=pulseMat)
   })
@@
 par(mar=c(4,5,2,2), mfrow=c(4,3))
-plot(sapply(res, function(x) x$IPI.summary[3])*1e6, ylab='Median interpulse interval\n(microsec)')
-abline(h=2*unique(paramMat$period), col='red')
+plot(sapply(res, function(x) x$IPI.summary[3])*1e6, ylab='Median interpulse interval\n(microsec)', log='y')
+abline(h=unique(paramMat$period), col='red')
 plot(sapply(res, function(x) x$IPI.sd)*1e6, ylab='Interpulse interval stdev\n(microsec)')
@@
 
-plot(sapply(res, function(x) x$PW.summary[3])*1e6, ylab='Median pulse width\n(microsec)')
+plot(sapply(res, function(x) x$PW.summary[3])*1e6, ylab='Median pulse width\n(microsec)', log='y')
 abline(h=2*unique(paramMat$dur0), col='red')
@@
 
+```{r fuckit}
+
+par(mfrow=c(3,3), mar=c(0,0,0,0))
+for (i in 1:10){#(length(res)/2)){
+  plot(rowMeans(res[[2*i-1]]$pulseMat), type='l', col='black', axes=FALSE, lty=1)
+  lines(rowMeans(res[[2*i]]$pulseMat), type='l', col='red', lty=1)
+}
+```
 ## Microstim is digital and highly configurable

### 2019-05-10 benchmarking/paper/draft2.Rnw
added analysis code for benchmarking

\documentclass{article}
\usepackage[margin=2cm]{geometry}
%\usepackage[superscript,biblabel]{cite}
\usepackage[super]{natbib}
\begin{document}
%\SweaveOpts{concordance=TRUE}

\title{Stimjim: open source hardware for precise electrical stimulation}
\author{}
\date{}
\maketitle

<<loadBenchmarkData, echo=FALSE, cache=TRUE>>=
library(parallel)

projectDirectory = '/home/x/Desktop/stimjim/benchmarking/'
setwd(projectDirectory)

BOM=openxlsx::read.xlsx("../stimjim_BOM.xlsx")
stimjimCost = ceiling(sum(BOM$Unit.cost*BOM$Quantity, na.rm=TRUE))

paramMat = read.csv(paste(projectDirectory,"paramMat.csv", sep=''))
colnames(paramMat) = c('mode0','mode1','period','duration','ch0amp0','ch1amp0','dur0', 'ch0amp1','ch1amp1','dur1')

vpau = 20/2^12 # volts per adc unit
fs = 2e6       # sampling rate of measurements

dataFiles = list.files(path=paste(projectDirectory,'data',sep=''), pattern="[.]bin", full.names = TRUE)

getPulseParams = function(f, thresh=25, fs, vpau){

  dat = matrix(readBin(f, what='integer', n=1e7, size=2, endian = 'big'),ncol=2,byrow=TRUE)

  apply(dat, 2, function(x) {  
    isPulse = 1.0*(x>thresh | x < -thresh)
    isPulse = mmand::erode(mmand::dilate(isPulse, rep(1,21)),rep(1,21))
    pulseIndices = which(isPulse>0)

    pulseStartIndices = pulseIndices[c(1,which(diff(pulseIndices) > 3)+1)]
    pulseEndIndices = pulseIndices[c(which(diff(pulseIndices) > 3), length(pulseIndices))]

    isStageHigh = 1.0*(x > thresh)
    isStageLow = 1.0*(x < -thresh)

    #plot(x[1:3000], type='l', ylim=c(190,220))
    #points(pulseIndices, x[pulseIndices], cex=0.3, col='red')
    #points(pulseStartIndices, x[pulseStartIndices], cex=1, col='blue')
    #points(pulseEndIndices, x[pulseEndIndices], cex=1, col='orange')

    IPI = diff(pulseStartIndices/fs)
    IPI.summary = summary(IPI)
    IPI.sd = sd(IPI)

    PW = (pulseEndIndices-pulseStartIndices)/fs
    PW.summary = summary(PW)
    PW.sd = sd(PW)

    sh.median = median(x[which(isStageHigh>0)])*vpau
    sh.sd = sd(x[which(mmand::erode(isStageHigh, rep(1,13))>0)])*vpau
    sl.median = median(x[which(isStageLow>0)])*vpau
    sl.sd = sd(x[which(mmand::erode(isStageLow, rep(1,13))>0)])*vpau

    pulseMat = sapply(pulseStartIndices, function(j) x[j+(-100:1000)])
    list(IPI.summary=IPI.summary, IPI.sd=IPI.sd, PW.summary=PW.summary, 
             PW.sd=PW.sd, sh.median=sh.median, sh.sd=sh.sd, 
             sl.median=sl.median, sl.sd=sl.sd, pulseMat=pulseMat)
  })
}

cl = makeCluster(detectCores()-1)
res = parLapply(cl, dataFiles, getPulseParams, fs=fs, vpau=vpau)
res = unlist(res, recursive = FALSE)
stopCluster(cl)
@

\section*{Abstract}
Electrical stimulation is a powerful tool to perturb and understand the function of neurons and neural circuits. Despite this, electronics that can provide precise current or voltage stimulation are expensive and closed-source. Here, we introduce stimjim, precise yet inexpensive (\$200) open-source hardware for electrical stimulation that combines both function generation and isolation. Stimjim provides microsecond temporal resolution with microampere or millivolt level precision on two isolated output channels. To show the utility of Stimjim, we demonstrate training of mice to perform intracranial self-stimulation (ICSS) for brain stimulation reward (BSR) in a novel head-fixed paradigm, and that interest in the reward can be tuned by adjusting either stimulation frequency or amplitude.

\section*{Introduction}
Electrical stimulation of neural tissue is an invaluable research tool. Over the past 150 years, it has enabled an understanding of the function of various brain regions by directly causing neurons in those regions to fire \cite{hitzig_uber_1870,ferrier_functions_1876,penfield_somatic_1937}. More recently, it has also found important clinicial applications in neurological disorders including Parkinson's disease\cite{deuschl_randomized_2006} and depression\cite{mayberg_deep_2005}. However, to date, the hardware enabling precise control over current and voltage stimulation remains generally expensive and closed source.

In contrast, there has been a recent push within the scientific community to produce open labware - open source hardware and software replacements for a variety of common laboratory tasks. Examples include Open Ephys\cite{siegle_open_2017}, Cyclops\cite{newman_optogenetic_2015}, OpenPCR\cite{noauthor_openpcr_nodate}, FlyPi\cite{chagas_100_2017}, Openspritzer\cite{forman_openspritzer:_2017}, and Audiomoth\cite{hill_audiomoth:_2018}. Many of these designs have been enabled by developments fueling the rapid growth of the maker community at large, including Arduino, Raspberry Pi, Teensy, openMV, pixyCam, open-source 3d printers like reprap.

\begin{table}
\small{
\begin{tabular}{|l|c|c|c|c|}
\hline
                                & Stimjim     & PulsePal     & STG400x (Multichannel Systems)    & PHM-15x (Med associates) \\
\hline
 Number of output channels      & 2           & 4            &  2-8                              & 2                        \\
 Current output mode            & X           &              &  X                                & X                        \\
 Voltage output mode            & X           & X            &  X                                &                          \\
 Outputs are electrically isolated           & X           &              &  X                                & X                        \\
 Onboard measurement            & X           &              &                                   & Optional                 \\
 Compliance voltage             & $\pm$ 13.7V & $\pm$ 15V    & $\pm$ 120V                        & $\pm$ 45V                \\
 Fastest pulse                  & 50us        & 10us         & 20us                              & 60us                     \\
 Cost (USD)                     & \$194       & \$200        & \$4131 (2 channels)               &                          \\
                                &             &              & \$7462  (4 channels)              &                          \\
                                &             &              & \$10327 (8 channels)              &                          \\
 Open source                    & X           & X            &                                   &                          \\
 \hline
\end{tabular}
}
\caption{\label{tab:stimulatorComparison} Comparison of Stimjim vs other related hardware. Cost was obtained as either the complete cost of the bill of materials (Stimjim and PulsePal) or quoted prices from manufacturers (STG400x and PHM-15x).}
\end{table}

\begin{figure}
\includegraphics[width=1\linewidth]{../../photo.png}
\caption{\textbf{Stimjim is a compact, configurable, and precise stimulator.}}
\label{fig:picture}
\end{figure}

\section*{Results}

To benchmark Stimjim, we varied the pulse frequency (from \Sexpr{round(1e6/max(paramMat$period),2)} Hz to \Sexpr{round(1e6/min(paramMat$period),2)} kHz), duration (from \Sexpr{min(paramMat$dur0)} to \Sexpr{max(paramMat$dur0)} us), and amplitude. We simultaneously recorded from both output channels on a National Instruments PCI-6110 card (12-bit ADC, sampling each channel at 2MHz).

<<benchmarkFigure, echo=FALSE,fig.height=8,fig.width=8,fig.cap="\\label{fig:benchmarking}">>=

par(mar=c(4,5,2,2), mfrow=c(4,3))

plotParams = list(
   list(j=1:30, column=3, label='Inter-pulse interval'),
   list(j=31:60, column=5, label='Inter-pulse interval'),
   list(j=61:90, column=7, label='Inter-pulse interval'))

for (i in 1:3){
  with(plotParams[[i]], {
    x = paramMat[j,column];

    plot(x, sapply(res[j], function(x) x$IPI.summary[3])*1e6, xlab=paste(label, "setting"), ylab=paste("Measured median",label), log='xy')
    plot(x, sapply(res[j], function(x) x$IPI.sd)*1e6,         xlab=paste(label, "setting"), ylab=paste("Measured median",label), log='xy')
    plot(x, sapply(res[j], function(x) x$IPI.sd)*1e6,         xlab=paste(label, "setting"), ylab=paste("Measured median",label), log='xy')
  })
}
@

To demonstrate the utility of Stimjim, we used it to train mice in a classical paradigm known as intra-cranial self stimulation (ICSS)\cite{olds_positive_1954}. In this assay, animals are implanted with electrodes (or more recently optical fibers) enabling activation of a pleasure/reward-related brain region\cite{carlezon_intracranial_2007}. Animals are then placed in a training paradigm in which they learn that a simple motor action (typically spinning a wheel or pressing a lever) causes direct activation of this brain region. Animals quickly learn the required action and are willing to repeat it for extended periods of time.

Here, we trained two mice in a head-fixed variant of ICSS, in which animals could lick a sensor in order to obtain brain stimulation reward (BSR). After animals had learned the licking behavior (usually within their first hour session), we varied the stimulation frequency and amplitude and assessed how it affected licking behavior. Both animals showed clear frequency- and amplitude-dependent responses, in which animals ceased licking when the rewarding stimulation was insufficiently intense (Fig. \ref{fig:icss}).

<<icssAnalysis, echo=FALSE, fig.cap="\\label{fig:icss}\\textbf{Stimjim enables measuring frequency- and amplitude-dependent responses in an intracranial self-stimulation paradigm.} Mice decreased their licking rates when the frequency and amplitude of the rewarding stimulation decreased. For each animal, three frequency sweeps (highest to lowest frequency, 1 minute per frequency) were performed in a 45-minute session. Amplitude sweeps were performed in the same way.", cache=TRUE, fig.height=5, fig.width=5>>=

icss.files = c("/home/x/Desktop/stimjim/benchmarking/2019-04-04_4309/20190404_115638.txt",
               "/home/x/Desktop/stimjim/benchmarking/2019-04-04_4309/20190404_143910.txt",
               "/home/x/Desktop/stimjim/benchmarking/2019-04-04_4476/20190404_130105.txt",
               "/home/x/Desktop/stimjim/benchmarking/2019-04-04_4476/20190404_181300.txt")

par(mar=c(4,4,1,1),mfrow=c(2,2), tck=0, mgp=c(1.5,0.1,0), oma=c(0,3,3,0))

for (i in 1:4){
  dat = read.csv(icss.files[i])
  names(dat) = c('time','c1','c2','pass','freq','amp', 'comment')

  x = mmand::dilate(dat$c1, kernel=rep(1,11))
  breaks = which(diff(dat$freq)!=0 | diff(dat$amp)!=0)
  amp = length(unique(dat$freq)) < length(unique(dat$amp))

  m = sapply(breaks, function(i) x[i+1000:11000])

  licks = apply(m-900,2, function(x) sum(c(x,0) > 0 & c(0,x) < 0))

  if (!amp){
    matplot(x=rev(sort(unique(dat$freq))), matrix(licks,nrow=15), type='p', ylim=c(0,30), bty='l', yaxt='n',
            xlab='Stimulation frequency (Hz)', ylab='licks per 50-second trial',lty=1,  pch=16, col=adjustcolor(1,0.5))
    lines(x=rev(sort(unique(dat$freq))), rowMeans(matrix(licks,nrow=15)), type='l', lwd=2)
  } else {
    matplot(x=rev(sort(unique(dat$amp))), matrix(licks,nrow=15), type='p', ylim=c(0,30), bty='l', yaxt='n',
            xlab='Stimulation amplitude (uA)', ylab='licks per 50-second trial',lty=1, pch=16, col=adjustcolor(1,0.5))
    lines(x=rev(sort(unique(dat$amp))), rowMeans(matrix(licks,nrow=15)), type='l', lwd=2)
  }
  axis(2,las=2)
}
mtext("Mouse 1", 2, adj=0.9,  outer=TRUE, cex=0.75)
mtext("Mouse 2", 2, adj=0.3,  outer=TRUE, cex=0.75)
mtext("Frequency Sweep", 3, adj=0.25,  outer=TRUE, cex=0.75)
mtext("Amplitude Sweep", 3, adj=0.85,  outer=TRUE, cex=0.75)

@

\section*{Conclusion}

\section*{Supplemental materials}

  - BOM
  - gerbers
  - schematic
  - link to bitbucket

\section*{References}
\bibliographystyle{unsrt}
\bibliography{stimjimBibliography}

\end{document}

### 2019-05-10 benchmarking/makeTestParameters.R
updated benchmarking code

@@ -15,13 +15,13 @@ ampScalings = seq(0,1000,length.out=30)
 pulseDurations = round(exp(seq(log(100),log(5000), length.out=30)))
 periods = round(exp(seq(log(500), log(5e5), length.out=30)))
 
-ampScaling=5; pulseDuration=100
+ampScaling=500; pulseDuration=100
 for (period in periods){
   params = c(mode0, mode1, period, duration, round(amp*ampScaling), pulseDuration, -round(amp*ampScaling), pulseDuration)
   paramMat = rbind(paramMat, params)        
 }
 
-ampScaling=5; period=10000;
+ampScaling=500; period=10000;
 for (pulseDuration in pulseDurations){
   params = c(mode0, mode1, period, duration, round(amp*ampScaling), pulseDuration, -round(amp*ampScaling), pulseDuration)
   paramMat = rbind(paramMat, params)        

### 2019-05-10 benchmarking/paper/draft2.Rnw
updates to benchmarking analysis

@@
 getPulseParams = function(f, thresh=25, fs, vpau){
-
   dat = matrix(readBin(f, what='integer', n=1e7, size=2, endian = 'big'),ncol=2,byrow=TRUE)
@@
 par(mar=c(4,5,2,2), mfrow=c(4,3))
+#matplot(sapply(res[1:60], function(x) x$pulseMat[,1]), col=c(1,2), type='l', lty=1)
+plot(paramMat[1:30,3], sapply(res[seq(1,60,by=2)], function(x) x$IPI.summary[3])*1e6, 
+  log='xy', main='median IPI vs target', cex=0.75, xlab='Target IPI', ylab='Median measured IPI')
+points(paramMat[1:30,3], sapply(res[seq(2,60,by=2)], function(x) x$IPI.summary[3])*1e6,
+  col='red', cex=0.75)
+abline(a=0,b=1)
 
-plotParams = list(
-   list(j=1:30, column=3, label='Inter-pulse interval'),
-   list(j=31:60, column=5, label='Inter-pulse interval'),
-   list(j=61:90, column=7, label='Inter-pulse interval'))
+#matplot(dendRites::offset(sapply(res[61:120], function(x) x$pulseMat[,1])), col=c(1,2), type='l', lty=1)
+plot(paramMat[1:30,3], sapply(res[seq(1,60,by=2)], function(x) x$IPI.sd)*1e6, 
+  log='x', main='IPI sd', cex=0.75, xlab='Target IPI', ylab='Measured IPI std. dev.')
+points(paramMat[1:30,3], sapply(res[seq(2,60,by=2)], function(x) x$IPI.sd)*1e6,
+  col='red', cex=0.75)
 
-for (i in 1:3){
-  with(plotParams[[i]], {
-    x = paramMat[j,column];
+plot(paramMat[1:30,3], sapply(res[seq(1,60,by=2)], function(x) diff(x$IPI.summary[c(1,6)]))*1e6, 
+  log='x', main='IPI range', cex=0.75, xlab='Target IPI', ylab='Range of measured IPIs')
+points(paramMat[1:30,3], sapply(res[seq(2,60,by=2)], function(x) diff(x$IPI.summary[c(1,6)]))*1e6,
+  col='red', cex=0.75)
 
-    plot(x, sapply(res[j], function(x) x$IPI.summary[3])*1e6, xlab=paste(label, "setting"), ylab=paste("Measured median",label), log='xy')
-    plot(x, sapply(res[j], function(x) x$IPI.sd)*1e6,         xlab=paste(label, "setting"), ylab=paste("Measured median",label), log='xy')
-    plot(x, sapply(res[j], function(x) x$IPI.sd)*1e6,         xlab=paste(label, "setting"), ylab=paste("Measured median",label), log='xy')
-  })
-}
+plot(paramMat[31:59,7], sapply(res[seq(61,118,by=2)], function(x) x$PW.summary[3])*1e6/2, 
+  log='xy', main='median PW vs target', ylab='measured median PW', cex=0.75)
+points(paramMat[31:59,7], sapply(res[seq(62,118,by=2)], function(x) x$PW.summary[3])*1e6/2,
+  col='red', cex=0.75)
+abline(a=0,b=1)
+
+plot(paramMat[31:59,7], sapply(res[seq(61,118,by=2)], function(x) x$PW.sd)*1e6/2, 
+  log='x', main='PW sd', ylab='PW sd', cex=0.75)
+points(paramMat[31:59,7], sapply(res[seq(62,118,by=2)], function(x) x$PW.sd)*1e6/2,
+  col='red', cex=0.75)
+
+plot(paramMat[31:59,7], sapply(res[seq(61,118,by=2)], function(x) diff(x$PW.summary[c(1,6)]))*1e6/2, 
+  log='x', main='PW range', ylab='PW range', cex=0.75)
+points(paramMat[31:59,7], sapply(res[seq(62,118,by=2)], function(x) diff(x$PW.summary[c(1,6)]))*1e6/2,
+  col='red', cex=0.75)
+
+plot(paramMat[61:90,5], sapply(res[seq(121,180,by=2)], function(x) x$sh.median*1e3), 
+  log='', main='amplitude vs target', xlab='Target amplitude (mV)', ylab='measured amplitude (mV)', cex=0.75)
+abline(a=0,b=1.0)
+points(paramMat[61:90,5], sapply(res[seq(122,180,by=2)], function(x) x$sh.median*1e3),
+  col='red', cex=0.75)
+
 @

### 2019-05-12 benchmarking/paper/draft2.Rnw
1) cleanup. 2) moved stimjim functions to library. 3) updates to draft

@@
  Fastest pulse                  & 50us        & 10us         & 20us                              & 60us                     \\
- Cost (USD)                     & \$194       & \$200        & \$4131 (2 channels)               &                          \\
-                                &             &              & \$7462  (4 channels)              &                          \\
+ Cost (USD)                     & \$194       & \$200        & \$4131 (2 channels)               &  \$6211 (plus \$4438     \\
+                                &             &              & \$7462  (4 channels)              &  for necessary software) \\
                                 &             &              & \$10327 (8 channels)              &                          \\
@@
 \section*{Results}
+We developed Stimjim to be an precision stimulus generator and isolator in a single package.
+Stimjim is based on the Teensy 3.5 microcontroller, which utilizes a 32-bit Arm Cortex-M4 processor running at 120MHz.
+Each stimulating channel includes a current source based on an improved Howland current pump, and a voltage source (an op-amp), driven by a 16-bit digital-to-analog converter (DAC).
+The final output of each channel is selected by a 4-way switch, such that either channel can be configured as a current output, voltage output, grounded, or disconnected (high-Z).
+Toe ensure the output is connected to a load and not either an open or short circuit, each channel also has a analog-to-digital converter able to read either the output voltage or the output current (via a sense resistor in series with the current output).
+
+Our design was made in Kicad, an open-source PCB CAD package.
+Schematic, layout, bill of materials, and build instructions are included in the supplement and in a git repository (https://bitbucket.org/natecermak/stimjim).
 
@@
 
-<<benchmarkFigure, echo=FALSE,fig.height=8,fig.width=8,fig.cap="\\label{fig:benchmarking}">>=
+<<benchmarkFigure, echo=FALSE,fig.height=7,fig.width=8,fig.cap="\\label{fig:benchmarking}">>=
 
-par(mar=c(4,5,2,2), mfrow=c(4,3))
-#matplot(sapply(res[1:60], function(x) x$pulseMat[,1]), col=c(1,2), type='l', lty=1)
+ex1 = sapply(dataFiles[seq(15,1,length.out = 6)], function(f){
+  dat = matrix(readBin(f, what='integer', n=1e7, size=2, endian = 'big'),ncol=2,byrow=TRUE)
+  fp = min(which(dat[,1] > 100))
+  dat[fp + (-1e3:3e4),1]*vpau
+})
+
+par(mar=c(4,4,1,1), mfrow=c(3,4), mgp=c(2,0.7,0))
+matplot(1e3/fs*1:nrow(ex1), dendRites::offset(ex1,15), col=1, type='l', lty=1, bty='n', yaxt='n', ylab='', xlab='time (ms)'); 
 plot(paramMat[1:30,3], sapply(res[seq(1,60,by=2)], function(x) x$IPI.summary[3])*1e6, 
@@
 
-#matplot(dendRites::offset(sapply(res[61:120], function(x) x$pulseMat[,1])), col=c(1,2), type='l', lty=1)
 plot(paramMat[1:30,3], sapply(res[seq(1,60,by=2)], function(x) x$IPI.sd)*1e6, 
@@
   col='red', cex=0.75)
+
+ex2 = sapply(dataFiles[seq(52,31,length.out=6)], function(f){
+  dat = matrix(readBin(f, what='integer', n=1e7, size=2, endian = 'big'),ncol=2,byrow=TRUE)
+  fp = min(which(dat[,1] > 100))
+  dat[fp + (-1e3:1e4),1]*vpau
+})
+
+matplot(1e3/fs*1:nrow(ex2), dendRites::offset(ex2,15), col=1, type='l', lty=1, bty='n', yaxt='n', ylab='', xlab='time (ms)');
 
@@
   col='red', cex=0.75)
+
+ex3 = sapply(dataFiles[seq(81,61,length.out=6)], function(f){
+  dat = matrix(readBin(f, what='integer', n=1e7, size=2, endian = 'big'),ncol=2,byrow=TRUE)
+  fp = min(which(dat[,1] > 10))
+  dat[fp + (-300:700),1]*vpau
+})
+
+matplot(1e3/fs*1:nrow(ex3), dendRites::offset(ex3,15), col=1, type='l', lty=1, bty='n', yaxt='n', ylab='', xlab='time (ms)');
 

### 2019-05-12 lib/stimjim/src/Stimjim.cpp

@@ -1,0 +1,188 @@
+#include "Stimjim.h"
+
+// ------------- SPI setup ------------------------------------- //
+//AD5752 DAC runs at max 30MHz, SPI mode 1 or 2
+SPISettings SpiSettingsDAC(10000000, MSBFIRST, SPI_MODE1); // Does not seem to work at 20MHz!
+//AD7321 settings - clock starts high, data latch on falling edge
+SPISettings SpiSettingsADC(5000000, MSBFIRST, SPI_MODE2);  // Can't run SPI clock faster than 1MHz without errors
+
+
+// ------------- Compensation for various DC offsets ---------- //
+int currentOffsets[2];
+int voltageOffsets[2];
+
+// --------- States of ADCs (set to read current or voltage?) ---//
+volatile bool adcSelectedInput[2];
+
+
+void setOutputMode(byte channel, byte mode) {
+  // mode 0-3 are: VOLTAGE, CURRENT, HIGH-Z, GROUND
+  digitalWrite((channel) ? OE0_1 : OE0_0, (0b00000001 & mode) ? HIGH : LOW);
+  digitalWrite((channel) ? OE1_1 : OE1_0, (0b00000010 & mode) ? HIGH : LOW);
+}
+
+
+
+void setADCrange(float range) {
+  SPI.beginTransaction(SpiSettingsADC);
+  for (int i = 0; i < 2; i++) {
+    int cs = (i == 0) ? CS1_0 : CS1_1; //CS1_x is ADC for channel x
+    digitalWrite(cs, LOW); //[15] = write, [13] = range register
+    if (range == 2.5) {
+      SPI.transfer16(0b1011000100000000); // write range register +-2.5V on both channels
+    } else if (range == 5) {
+      SPI.transfer16(0b1010100010000000); // write range register +-5V on both channels)
+    } else {
+      SPI.transfer16(32768 + 8192);       // write range register, +-10V on both channels
+    }
+    digitalWrite(cs, HIGH);
+  }
+  SPI.endTransaction();
+}
+
+void setupDACs() {
+  SPI.beginTransaction(SpiSettingsDAC);
+  for (int i = 0; i < 2; i++) {
+    int cs = (i == 0) ? CS0_0 : CS0_1; //CS0_x is DAC for channel x
[…]

### 2019-05-12 lib/stimjim/src/Stimjim.h

@@ -1,0 +1,59 @@
+/* Stimjim library (c) Nathan Cermak 2019. */
+
+#ifndef __STIMJIM_H__
+#define __STIMJIM_H__
+
+
+#if ARDUINO >= 100
+#include <Arduino.h>
+#else
+#include <WProgram.h>
+#endif
+#include <SPI.h>
+
+#define MICROAMPS_PER_DAC 0.1017            // 20V * (1/(3000 V/A)) / 2^16 = 0.1uA / DAC unit
+#define MICROAMPS_PER_ADC 0.85              //  (1/(100*(1+49.9k/1.8k) V/A)) * 20V / 2^13 = 0.85
+#define MILLIVOLTS_PER_DAC 0.442            // 20V / 2^16 * gain of 1.45 ~= 0.44 mV / DAC unit
+#define MILLIVOLTS_PER_ADC 2.44             // 20V / 2^13 = 2.44 mV / ADC unit
+
+#define OE0_0 6
+#define OE1_0 5
+#define OE0_1 1
+#define OE1_1 0
+#define CS0_0 9
+#define CS1_0 7
+#define CS0_1 3
+#define CS1_1 2
+#define NLDAC_1 4
+#define NLDAC_0 10
+#define IN0 22
+#define IN1 23
+#define LED0 21
+#define LED1 20
+#define MISO_0 8
+#define MISO_1 12
+
+
+
+
+
+// ------------ Function prototypes --------------------------- //
+void setOutputMode(byte channel, byte mode);
+void setADCrange(float range);
+void setupDACs();
+void writeToDacs(int16_t amp0, int16_t amp1);
+void writeToDac(byte channel, int16_t amp);
+int readADC(byte channel, byte line);
+void getCurrentOffsets();
+void getVoltageOffsets();
+
+
+// ------------- Compensation for various DC offsets ---------- //
+extern int currentOffsets[2];
+extern int voltageOffsets[2];
+
+// --------- States of ADCs (set to read current or voltage?) ---//
+extern volatile bool adcSelectedInput[2];
+
+
+#endif /* ! defined __STIMJIM_H__ */

### 2019-05-12 benchmarking/makeTestParameters.R
changed sweep ranges for benchmarking

@@ -11,9 +11,9 @@ paramMat = NULL
 mode0 = 0
 mode1 = 1
 amp = c(10,1)
+periods = round(exp(seq(log(250), log(4e5), length.out=30)))
+pulseDurations = round(exp(seq(log(20),log(5000), length.out=30)))
 ampScalings = seq(0,1000,length.out=30)
-pulseDurations = round(exp(seq(log(100),log(5000), length.out=30)))
-periods = round(exp(seq(log(500), log(5e5), length.out=30)))
 
 ampScaling=500; pulseDuration=100
 for (period in periods){

### 2019-05-12 benchmarking/paper/draft2.Rnw
fixed leading space in paramStringFile, updated analysis

@@
 
-plot(paramMat[61:90,5], sapply(res[seq(121,180,by=2)], function(x) x$sh.median*1e3), 
-  log='', main='amplitude vs target', xlab='Target amplitude (mV)', ylab='measured amplitude (mV)', cex=0.75)
+x = c(paramMat[61:90,5], -paramMat[61:90,5])
+y = t(sapply(res[seq(121,180,by=2)], function(x) c(x$sh.median,x$sl.median))*1e3)
+plot(x,y, main='Voltage accuracy', xlab='Target amplitude (mV)', ylab='measured amplitude (mV)', cex=0.75)
 abline(a=0,b=1.0)
-points(paramMat[61:90,5], sapply(res[seq(122,180,by=2)], function(x) x$sh.median*1e3),
-  col='red', cex=0.75)
+
+x = c(paramMat[61:90,6], -paramMat[61:90,6])
+y = t(sapply(res[seq(122,180,by=2)], function(x) c(x$sh.median,x$sl.median))/9750*1e6)
+plot(x,y, main='Current accuracy', xlab='Target amplitude (uA)', ylab='measured amplitude (uA)', cex=0.75, col='red')
+abline(a=0,b=1)
 

### 2019-05-12 benchmarking/makeTestParameters.R

@@ -38,7 +38,7 @@ par2string = function(p){
 }
 
 stringVector = apply(paramMat,1,par2string)
-cat(stringVector, file=paste(projectDirectory,"benchmarking/paramStringFile.txt", sep=''))
+cat(stringVector, file=paste(projectDirectory,"benchmarking/paramStringFile.txt", sep=''), sep='')
 write.csv(paramMat, paste(projectDirectory,"benchmarking/paramMat.csv", sep=''),row.names = FALSE)
 
 

### 2019-05-12 benchmarking/makeTestParameters.R
changed benchmarking sweep range for amplitude to start at 20mV or uA.

@@ -13,7 +13,7 @@ mode1 = 1
 amp = c(10,1)
 periods = round(exp(seq(log(250), log(4e5), length.out=30)))
 pulseDurations = round(exp(seq(log(20),log(5000), length.out=30)))
-ampScalings = seq(0,1000,length.out=30)
+ampScalings = seq(20,1000,length.out=30)
 
 ampScaling=500; pulseDuration=100
 for (period in periods){

### 2019-05-12 benchmarking/paper/draft2.Rnw
fixed wrong gain in code from old (no longer correct) resistor values

@@
 
-x = c(paramMat[61:90,6], -paramMat[61:90,6])
-y = t(sapply(res[seq(122,180,by=2)], function(x) c(x$sh.median,x$sl.median))/9750*1e6)
-plot(x,y, main='Current accuracy', xlab='Target amplitude (uA)', ylab='measured amplitude (uA)', cex=0.75, col='red')
+x2 = c(paramMat[61:90,6], -paramMat[61:90,6])
+y2 = t(sapply(res[seq(122,180,by=2)], function(x) c(x$sh.median,x$sl.median))/9750*1e6)
+plot(x2,y2, main='Current accuracy', xlab='Target amplitude (uA)', ylab='measured amplitude (uA)', cex=0.75, col='red')
 abline(a=0,b=1)
+
+plot(x, as.vector(y)-x, pch=16, xlab='', ylab='')
+par(new=TRUE)
+plot(x2, as.vector(y2)-x2, pch=16, axes=F, col='red', xlab='',ylab='')
+axis(1, col='red',line=2); 
+axis(4, col='red')
 

### 2019-05-12 lib/stimjim/src/Stimjim.h

@@ -13,7 +13,7 @@
 
 #define MICROAMPS_PER_DAC 0.1017            // 20V * (1/(3000 V/A)) / 2^16 = 0.1uA / DAC unit
 #define MICROAMPS_PER_ADC 0.85              //  (1/(100*(1+49.9k/1.8k) V/A)) * 20V / 2^13 = 0.85
-#define MILLIVOLTS_PER_DAC 0.442            // 20V / 2^16 * gain of 1.45 ~= 0.44 mV / DAC unit
+#define MILLIVOLTS_PER_DAC 0.458             // 20V / 2^16 * gain of 1.5 ~= 0.44 mV / DAC unit
 #define MILLIVOLTS_PER_ADC 2.44             // 20V / 2^13 = 2.44 mV / ADC unit
 
 #define OE0_0 6

### 2019-05-12 benchmarking/paper/draft2.Rnw
got pulse widths calibrated right.

@@
 res = parLapply(cl, dataFiles, getPulseParams, fs=fs, vpau=vpau)
+#res = list(); for (i in 1:length(dataFiles)) res[[i]] = getPulseParams(dataFiles[i], fs=fs, vpau=vpau)
+
 res = unlist(res, recursive = FALSE)
@@
 \small{
-\begin{tabular}{|l|c|c|c|c|}
+\begin{tabular}{|l|c|c|c|c|c|}
 \hline
-                                & Stimjim     & PulsePal     & STG400x (Multichannel Systems)    & PHM-15x (Med associates) \\
+                                &             &              & STG400x                           & PHM-15x                  &  \\
+                                & Stimjim     & PulsePal     & (Multichannel Systems)            & (Med associates)         & ISO-O1B (NPI) \\
 \hline
- Number of output channels      & 2           & 4            &  2-8                              & 2                        \\
- Current output mode            & X           &              &  X                                & X                        \\
- Voltage output mode            & X           & X            &  X                                &                          \\
- Outputs are electrically isolated           & X           &              &  X                                & X                        \\
- Onboard measurement            & X           &              &                                   & Optional                 \\
- Compliance voltage             & $\pm$ 13.7V & $\pm$ 15V    & $\pm$ 120V                        & $\pm$ 45V                \\
- Fastest pulse                  & 50us        & 10us         & 20us                              & 60us                     \\
- Cost (USD)                     & \$194       & \$200        & \$4131 (2 channels)               &  \$6211 (plus \$4438     \\
-                                &             &              & \$7462  (4 channels)              &  for necessary software) \\
-                                &             &              & \$10327 (8 channels)              &                          \\
- Open source                    & X           & X            &                                   &                          \\
+ Number of output channels      & 2           & 4            &  2-8                              & 2                        & 1             \\ 
+ Stimulus generator included?   & X           & X            &  X                                & X                        &               \\  
+ Current output mode            & X           &              &  X                                & X                        & X                        \\
+ Voltage output mode            & X           & X            &  X                                &                          & X                        \\
+ Outputs electrically isolated  & X           &              &  X                                & X                        & X                        \\
+ Onboard measurement            & X           &              &                                   & Optional                 & X (compliance warning)   \\
+ Compliance voltage             & $\pm$ 13.7V & $\pm$ 15V    & $\pm$ 120V                        & $\pm$ 45V                & $\pm$ 100V     \\
+ Fastest pulse                  & 50us        & 10us         & 20us                              & 60us                     & 10us           \\
+ Cost (USD)                     & \$194       & \$200        & \$4131 (2 channels)               &  \$6211 (plus \$4438     & \$1708         \\
+                                &             &              & \$7462  (4 channels)              &  for necessary software) &                \\
+                                &             &              & \$10327 (8 channels)              &                          &                \\
+ Open source                    & X           & X            &                                   &                          &                \\
  \hline
@@
   fp = min(which(dat[,1] > 10))
-  dat[fp + (-300:700),1]*vpau
+  dat[fp + (-200:700),1]*vpau
 })

### 2019-05-12 benchmarking/makeTestParameters.R

@@ -12,7 +12,7 @@ mode0 = 0
 mode1 = 1
 amp = c(10,1)
 periods = round(exp(seq(log(250), log(4e5), length.out=30)))
-pulseDurations = round(exp(seq(log(20),log(5000), length.out=30)))
+pulseDurations = round(exp(seq(log(25),log(5000), length.out=30)))
 ampScalings = seq(20,1000,length.out=30)
 
 ampScaling=500; pulseDuration=100

### 2019-05-12 lib/stimjim/src/Stimjim.cpp

@@ -17,8 +17,8 @@ volatile bool adcSelectedInput[2];
 
 void setOutputMode(byte channel, byte mode) {
   // mode 0-3 are: VOLTAGE, CURRENT, HIGH-Z, GROUND
-  digitalWrite((channel) ? OE0_1 : OE0_0, (0b00000001 & mode) ? HIGH : LOW);
-  digitalWrite((channel) ? OE1_1 : OE1_0, (0b00000010 & mode) ? HIGH : LOW);
+  digitalWriteFast((channel) ? OE0_1 : OE0_0, (0b00000001 & mode) ? HIGH : LOW);
+  digitalWriteFast((channel) ? OE1_1 : OE1_0, (0b00000010 & mode) ? HIGH : LOW);
 }
 
 

### 2019-05-12 benchmarking/makeTestParameters.R
changed sweep parameters again.

@@ -11,8 +11,8 @@ paramMat = NULL
 mode0 = 0
 mode1 = 1
 amp = c(10,1)
-periods = round(exp(seq(log(250), log(4e5), length.out=30)))
-pulseDurations = round(exp(seq(log(25),log(5000), length.out=30)))
+periods = round(exp(seq(log(250), log(5e5), length.out=30)))
+pulseDurations = round(exp(seq(log(25),log(4000), length.out=30)))
 ampScalings = seq(20,1000,length.out=30)
 
 ampScaling=500; pulseDuration=100

### 2019-05-14 benchmarking/paper/draft2.Rnw
updates to analysis code

@@
-\documentclass{article}
+\documentclass[twocolumn]{article}
 \usepackage[margin=2cm]{geometry}
@@
 
-\title{Stimjim: open source hardware for precise electrical stimulation}
+\title{Stimjim is open source hardware for precise electrical stimulation}
 \author{}
@@
 
-\begin{table}
+\section*{Results}
+\textbf{Design.} We developed Stimjim to be an precision stimulus generator and isolator in a single package.
+Stimjim is based on the Teensy 3.5 microcontroller, which utilizes a 32-bit Arm Cortex-M4 processor running at 120MHz.
+Each stimulating channel includes a current source based on an improved Howland current pump, and a voltage source (an op-amp), driven by a 16-bit digital-to-analog converter (DAC).
+The final output of each channel is selected by a 4-way switch, such that either channel can be configured as a current output, voltage output, grounded, or disconnected (high-Z).
+To ensure the output is connected to a load and not either an open or short circuit, each channel also has a analog-to-digital converter able to read either the output voltage or the output current (via a sense resistor in series with the current output).
+Our design was made in Kicad, an open-source PCB CAD package.
+Schematic, layout, bill of materials, and build instructions are included as supplemental materials and are also available in a git repository (https://bitbucket.org/natecermak/stimjim).
+
+\begin{figure}[h]
+\begin{centering}
+\includegraphics[width=0.85\linewidth]{../../photo.png}
+\caption{\textbf{Stimjim is a compact, configurable, and precise stimulator.}}
+\label{fig:picture}
+\end{centering}
+\end{figure}
+
+Stimjim's design is significantly less expensive than other commercial alternatives (Table 1).
+Its specifications are otherwise similar or better, with the notable exception that its compliance voltage is lower than other stimulators.
+This limits the load resistance that Stimjim can drive (for a given resistance R, Stimjim cannot output a current larger than $\frac{13.7V}{R}$).
+While this may be a problem for electrodes with very high impedances (e.g.
+sharp intracellular or patch electrodes), it is unlikely to be a problem for implanted stimulating electrodes.
+
+\begin{table*}[t]
 \small{
@@
 \hline
-                                &             &              & STG400x                           & PHM-15x                  &  \\
-                                & Stimjim     & PulsePal     & (Multichannel Systems)            & (Med associates)         & ISO-O1B (NPI) \\
+                                &               &                  & STG400x                           & PHM-15x                  & ISO-01B       \\
+                                & Stimjim       & PulsePal 2       & (Multichannel Systems)            & (Med Associates)         & (NPI)         \\
 \hline
- Number of output channels      & 2           & 4            &  2-8                              & 2                        & 1             \\ 
- Stimulus generator included?   & X           & X            &  X                                & X                        &               \\  
- Current output mode            & X           &              &  X                                & X                        & X                        \\
- Voltage output mode            & X           & X            &  X                                &                          & X                        \\
- Outputs electrically isolated  & X           &              &  X                                & X                        & X                        \\
- Onboard measurement            & X           &              &                                   & Optional                 & X (compliance warning)   \\
- Compliance voltage             & $\pm$ 13.7V & $\pm$ 15V    & $\pm$ 120V                        & $\pm$ 45V                & $\pm$ 100V     \\
- Fastest pulse                  & 50us        & 10us         & 20us                              & 60us                     & 10us           \\
- Cost (USD)                     & \$194       & \$200        & \$4131 (2 channels)               &  \$6211 (plus \$4438     & \$1708         \\
-                                &             &              & \$7462  (4 channels)              &  for necessary software) &                \\
-                                &             &              & \$10327 (8 channels)              &                          &                \\
- Open source                    & X           & X            &                                   &                          &                \\
+ Output channels                & 2             & 4                &  2-8                              & 2                        & 1             \\ 
+ Stimulus generator included    & Y             & Y                &  Y                                & Y                        & N             \\  
+ Current output mode            & Y             & N                &  Y                                & Y                        & Y                        \\
+ Voltage output mode            & Y             & Y                &  Y                                & N                        & Y                        \\
+ Outputs electrically isolated  & Y             & N                &  Y                                & Y                        & Y                        \\
+ Onboard measurement            & Y             & N                &  N                                & Optional                 & Y*             \\
+ Compliance voltage             & $\pm$ 15V (voltage) & $\pm$ 10V  & $\pm$ 120V                        & $\pm$ 45V                & $\pm$ 100V     \\
+                                & $\pm$ 13.7V (current)    &       &                                   &                          &                \\
+ Fastest pulse                  & 30$\mu$s      & 100 $\mu$s       & 20 $\mu$s                         & 60  $\mu$s               & 10 $\mu$s      \\
+ Cost (USD)                     & \$194 (parts) & \$264 (parts)    & \$4131 (2 channels)               &  \$6211 (+ \$4438     & \$1708         \\
+                                &               & \$695 (assembled,& \$7462  (4 channels)              &  for software)           &                \\
+                                &               & from Sanworks)   & \$10327 (8 channels)              &                          &                \\
+ Open source                    & Y             & Y                &  N                                &  N                       & N              \\
  \hline
@@
 }
-\caption{\label{tab:stimulatorComparison} Comparison of Stimjim vs other related hardware.
-Cost was obtained as either the complete cost of the bill of materials (Stimjim and PulsePal) or quoted prices from manufacturers (STG400x and PHM-15x).}
-\end{table}
+\caption{\label{tab:stimulatorComparison} \textbf{Comparison of Stimjim vs other related hardware.} Cost was obtained as either the complete cost of the bill of materials (Stimjim and PulsePal) or quoted prices from distributors (STG400x, PHM-15x, ISO-01B).
+*ISO-01B has a compliance warning sound if the load resistance is too high, but does not report actual current measurements.}
+\end{table*}
 
-\begin{figure}
-\includegraphics[width=1\linewidth]{../../photo.png}
-\caption{\textbf{Stimjim is a compact, configurable, and precise stimulator.}}
-\label{fig:picture}
-\end{figure}
+\textbf{Benchmarking.} To benchmark Stimjim, we varied the pulse frequency (from \Sexpr{round(1e6/max(paramMat$period),2)} to \Sexpr{round(1e6/min(paramMat$period),2)} Hz), duration (from \Sexpr{min(paramMat$dur0)} to \Sexpr{max(paramMat$dur0)} $\mu$s), and amplitude while simultaneously recording from both output channels on a National Instruments PCI-6110 card (12-bit ADC, sampling each channel at 2MHz).
+We set one channel to voltage mode and the other channel to current mode with a 9.75 k$\Omega$ resistor connected to the output.
 
-\section*{Results}
-We developed Stimjim to be an precision stimulus generator and isolator in a single package.
-Stimjim is based on the Teensy 3.5 microcontroller, which utilizes a 32-bit Arm Cortex-M4 processor running at 120MHz.
-Each stimulating channel includes a current source based on an improved Howland current pump, and a voltage source (an op-amp), driven by a 16-bit digital-to-analog converter (DAC).
-The final output of each channel is selected by a 4-way switch, such that either channel can be configured as a current output, voltage output, grounded, or disconnected (high-Z).
-Toe ensure the output is connected to a load and not either an open or short circuit, each channel also has a analog-to-digital converter able to read either the output voltage or the output current (via a sense resistor in series with the current output).
+Stimjim proved exceptionally accurate, providing microsecond temporal resolution and millivolt- and microampere- amplitude resolution.
+Across the tested range of stimulation frequencies, Stimjim provided accurate and highly consistent inter-pulse intervals (IPIs; Fig.
+XA-C).
+The standard deviation of interpulse intervals within a 1-second pulse train was always below one microsecond.
+Stimjim also produced pulse widths that were accurate and minimally variable (Fig XD-F).
+Finally, we found Stimjim produced accurate voltage and current amplitudes.
 
-Our design was made in Kicad, an open-source PCB CAD package.
-Schematic, layout, bill of materials, and build instructions are included in the supplement and in a git repository (https://bitbucket.org/natecermak/stimjim).
+<<benchmarkFigure, echo=FALSE,fig.height=5.5,fig.width=5,fig.cap="\\label{fig:benchmarking} \\textbf{Electrical benchmarks show Stimjim provides microsecond temporal precision and millivolt- and microampere-scale amplitude precision.}", fig.align='center', out.width='1.1\\linewidth', fig.env="figure">>=
 
-To benchmark Stimjim, we varied the pulse frequency (from \Sexpr{round(1e6/max(paramMat$period),2)} Hz to \Sexpr{round(1e6/min(paramMat$period),2)} kHz), duration (from \Sexpr{min(paramMat$dur0)} to \Sexpr{max(paramMat$dur0)} us), and amplitude.
-We simultaneously recorded from both output channels on a National Instruments PCI-6110 card (12-bit ADC, sampling each channel at 2MHz).
-
-<<benchmarkFigure, echo=FALSE,fig.height=7,fig.width=8,fig.cap="\\label{fig:benchmarking}">>=
+options(scipen=999)
 
@@
 
-par(mar=c(4,4,1,1), mfrow=c(3,4), mgp=c(2,0.7,0))
-matplot(1e3/fs*1:nrow(ex1), dendRites::offset(ex1,15), col=1, type='l', lty=1, bty='n', yaxt='n', ylab='', xlab='time (ms)'); 
-plot(paramMat[1:30,3], sapply(res[seq(1,60,by=2)], function(x) x$IPI.summary[3])*1e6, 
-  log='xy', main='median IPI vs target', cex=0.75, xlab='Target IPI', ylab='Median measured IPI')
-points(paramMat[1:30,3], sapply(res[seq(2,60,by=2)], function(x) x$IPI.summary[3])*1e6,
+layout(matrix(1:9,3,3,byrow=TRUE), widths=c(0.8,1,1))
+par(mar=c(4,1,2,1), mgp=c(2,0.3,0), cex.main=1.05, tck=-0.01)
+matplot(1e3/fs*1:nrow(ex1), dendRites::offset(ex1,15), col=1, type='l', lty=1, bty='n', yaxt='n', ylab='', xlab='time (ms)',
+        main = "Varying frequency\n(Inter-pulse interval, IPI)"); 
+par(mar=c(4,4,2,1))
+plot(paramMat[1:30,3]/1e3, sapply(res[seq(1,60,by=2)], function(x) x$IPI.summary[3])*1e3, yaxt='n',
+  log='xy', main='', cex=0.75, xlab='Target IPI (ms)', ylab='Measured IPI (ms)')
+axis(2,las=2)
+points(paramMat[1:30,3]/1e3, sapply(res[seq(2,60,by=2)], function(x) x$IPI.summary[3])*1e3, 
   col='red', cex=0.75)
@@
 
-plot(paramMat[1:30,3], sapply(res[seq(1,60,by=2)], function(x) x$IPI.sd)*1e6, 
-  log='x', main='IPI sd', cex=0.75, xlab='Target IPI', ylab='Measured IPI std. dev.')
-points(paramMat[1:30,3], sapply(res[seq(2,60,by=2)], function(x) x$IPI.sd)*1e6,
+plot(paramMat[1:30,3]/1e3, sapply(res[seq(1,60,by=2)], function(x) x$IPI.sd)*1e6, yaxt='n',
+  log='x', main='', cex=0.75, xlab='Target IPI (ms)', ylab='IPI standard deviation (us)')
+points(paramMat[1:30,3]/1e3, sapply(res[seq(2,60,by=2)], function(x) x$IPI.sd)*1e6,
   col='red', cex=0.75)
-
-plot(paramMat[1:30,3], sapply(res[seq(1,60,by=2)], function(x) diff(x$IPI.summary[c(1,6)]))*1e6, 
-  log='x', main='IPI range', cex=0.75, xlab='Target IPI', ylab='Range of measured IPIs')
-points(paramMat[1:30,3], sapply(res[seq(2,60,by=2)], function(x) diff(x$IPI.summary[c(1,6)]))*1e6,
-  col='red', cex=0.75)
+axis(2,las=2)
+# 
+# plot(paramMat[1:30,3], sapply(res[seq(1,60,by=2)], function(x) diff(x$IPI.summary[c(1,6)]))*1e6, 
+#   log='x', main='IPI range', cex=0.75, xlab='Target IPI', ylab='Range of measured IPIs')
+# points(paramMat[1:30,3], sapply(res[seq(2,60,by=2)], function(x) diff(x$IPI.summary[c(1,6)]))*1e6,
+#   col='red', cex=0.75)
 
@@
   fp = min(which(dat[,1] > 100))
-  dat[fp + (-1e3:1e4),1]*vpau
+  dat[fp + (-1e3:5e3),1]*vpau
 })
-
-matplot(1e3/fs*1:nrow(ex2), dendRites::offset(ex2,15), col=1, type='l', lty=1, bty='n', yaxt='n', ylab='', xlab='time (ms)');
-
-plot(paramMat[31:59,7], sapply(res[seq(61,118,by=2)], function(x) x$PW.summary[3])*1e6/2, 
-  log='xy', main='median PW vs target', ylab='measured median PW', cex=0.75)
+par(mar=c(4,1,2,1))
+matplot(1e3/fs*1:nrow(ex2), dendRites::offset(ex2,15), col=1, type='l', lty=1, bty='n', yaxt='n', ylab='', xlab='time (ms)',
+                main = "Varying pulse width (PW)"); 
+par(mar=c(4,4,2,1))
+plot(paramMat[31:59,7], sapply(res[seq(61,118,by=2)], function(x) x$PW.summary[3])*1e6/2,  yaxt='n',
+  log='xy', main='', xlab='Target PW (us)', ylab='Measured PW (us)', cex=0.75)
 points(paramMat[31:59,7], sapply(res[seq(62,118,by=2)], function(x) x$PW.summary[3])*1e6/2,
@@
 abline(a=0,b=1)
+axis(2,las=2)
 
-plot(paramMat[31:59,7], sapply(res[seq(61,118,by=2)], function(x) x$PW.sd)*1e6/2, 
-  log='x', main='PW sd', ylab='PW sd', cex=0.75)
+plot(paramMat[31:59,7], sapply(res[seq(61,118,by=2)], function(x) x$PW.sd)*1e6/2, yaxt='n',
+  log='x', main='', xlab='Target PW (us)', ylab='PW standard deviation (us)', cex=0.75)
 points(paramMat[31:59,7], sapply(res[seq(62,118,by=2)], function(x) x$PW.sd)*1e6/2,
   col='red', cex=0.75)
-
-plot(paramMat[31:59,7], sapply(res[seq(61,118,by=2)], function(x) diff(x$PW.summary[c(1,6)]))*1e6/2, 
-  log='x', main='PW range', ylab='PW range', cex=0.75)
-points(paramMat[31:59,7], sapply(res[seq(62,118,by=2)], function(x) diff(x$PW.summary[c(1,6)]))*1e6/2,
-  col='red', cex=0.75)
+axis(2,las=2)
+# plot(paramMat[31:59,7], sapply(res[seq(61,118,by=2)], function(x) diff(x$PW.summary[c(1,6)]))*1e6/2, 
+#   log='x', main='PW range', ylab='PW range', cex=0.75)
+# points(paramMat[31:59,7], sapply(res[seq(62,118,by=2)], function(x) diff(x$PW.summary[c(1,6)]))*1e6/2,
+#   col='red', cex=0.75)
 
@@
 })
-
-matplot(1e3/fs*1:nrow(ex3), dendRites::offset(ex3,15), col=1, type='l', lty=1, bty='n', yaxt='n', ylab='', xlab='time (ms)');
-
-x = c(paramMat[61:90,5], -paramMat[61:90,5])
-y = t(sapply(res[seq(121,180,by=2)], function(x) c(x$sh.median,x$sl.median))*1e3)
-plot(x,y, main='Voltage accuracy', xlab='Target amplitude (mV)', ylab='measured amplitude (mV)', cex=0.75)
-abline(a=0,b=1.0)
+par(mar=c(4,1,2,1))
+matplot(1e3/fs*1:nrow(ex3), dendRites::offset(ex3,15), col=1, type='l', lty=1, bty='n', yaxt='n', ylab='', xlab='time (ms)',
+                main = "Varying amplitude");
+par(mar=c(4,4,2,1))
+x = c(paramMat[61:89,5], -paramMat[61:89,5])
+y = t(sapply(res[seq(121,178,by=2)], function(x) c(x$sh.median,x$sl.median))*1e3)
+plot(x/1000,y-x, main='Voltage accuracy', xlab='Target amplitude (V)', ylab='Measured error (mV)', cex=0.75, yaxt='n', ylim=c(-30,30))
+axis(2,las=2)
 
@@
 y2 = t(sapply(res[seq(122,180,by=2)], function(x) c(x$sh.median,x$sl.median))/9750*1e6)
-plot(x2,y2, main='Current accuracy', xlab='Target amplitude (uA)', ylab='measured amplitude (uA)', cex=0.75, col='red')
-abline(a=0,b=1)
+plot(x2/1000,y2-x2, main='Current accuracy', xlab='Target amplitude (mA)', ylab='Measured error (uA)', cex=0.75, col='red', yaxt='n',ylim=c(-5,5))
+axis(2,las=2)
 
-plot(x, as.vector(y)-x, pch=16, xlab='', ylab='')
-par(new=TRUE)
-plot(x2, as.vector(y2)-x2, pch=16, axes=F, col='red', xlab='',ylab='')
-axis(1, col='red',line=2); 
-axis(4, col='red')
+# plot(x, as.vector(y)-x, pch=16, xlab='', ylab='')
+# par(new=TRUE)
+# plot(x2, as.vector(y2)-x2, pch=16, axes=F, col='red', xlab='',ylab='')
+# axis(1, col='red',line=2); 
+# axis(4, col='red')
 
@@
 
-To demonstrate the utility of Stimjim, we used it to train mice in a classical paradigm known as intra-cranial self stimulation (ICSS)\cite{olds_positive_1954}.
+\textbf{In vivo stimulation.} To demonstrate the utility of Stimjim, we used it to train mice in a classical paradigm known as intra-cranial self stimulation (ICSS)\cite{olds_positive_1954}.
 In this assay, animals are implanted with electrodes (or more recently optical fibers) enabling activation of a pleasure/reward-related brain region\cite{carlezon_intracranial_2007}.
@@
 For each animal, three frequency sweeps (highest to lowest frequency, 1 minute per frequency) were performed in a 45-minute session.
-Amplitude sweeps were performed in the same way.", cache=TRUE, fig.height=5, fig.width=5>>=
+Amplitude sweeps were performed in the same way.", cache=TRUE, fig.height=5, fig.width=5, fig.align='center'>>=
 
@@
 
-par(mar=c(4,4,1,1),mfrow=c(2,2), tck=0, mgp=c(1.5,0.1,0), oma=c(0,3,3,0))
+par(mar=c(4,4,1,1),mfrow=c(2,2), tck=0, mgp=c(1.5,0.1,0), oma=c(0,1,3,0))
 
@@
 
+\section*{Acknowledgments}
+We thank the Schiller lab for discussions and help with preliminary testing. This work was supported in part by the Zuckerman STEM Leadership Program (NC).
+
+\section*{Methods}
+\textbf{Electrode implantation}.
+Monopolar electrodes (Plastics1, MS303/2-AIU/SPC) were implanted above the medial forebrain bundle according to the protocol in reference \cite{carlezon_intracranial_2007}.
+Additionally, a 3D-printed headpost was affixed to the animal's skull by dental cement to enable head fixation.
+All animal procedures were in accordance with guidelines established bythe NIH on the care and use of animals in research and were confirmed by the Technion Institutional Animal Care and Use Committee (IL-012-01-18, valid until 10/4/2022).
+
 \section*{Supplemental materials}
+BOM, gerbers, schematic, link to bitbucket
 
-  - BOM
-  - gerbers
-  - schematic
-  - link to bitbucket
-
-\section*{References}
-\bibliographystyle{unsrt}
+\bibliographystyle{nat}
 \bibliography{stimjimBibliography}
 
+\begin{table}
+\small{
+\begin{tabular}{|l|c|c|}
+\hline
+                                            & Value     & Units     \\
+\hline
+ Slew rate (voltage mode)                    & 4.1         & V/us     \\
+ Slew rate (current mode)                    & 0.8         & V/us     \\
+ Output voltage noise (voltage mode)$^a$     & 1           & mV rms    \\
+                                             & 5           & mVpp      \\
+ Output voltage noise (current mode)$^a$     & 0.6         & mV rms     \\
+                                             & 3           & mVpp      \\
+ Voltage output step size (voltage mode)$^b$ & 0.45        & V         \\
+ Current output step size (current mode)$^b$ & 0.1         & uA        \\ 
+ Trigger latency                             & 23          & us        \\
+ \hline
+\end{tabular}
+}
+\caption{\label{tab:stimjimSpecs} \textbf{Stimjim electrical characteristics.} $^a$Measured on Tektronix TDS210 digital oscilloscope, full bandwidth (60MHz), 25ms segments..
+Note that this is the noise only when the pulse is delivered; at all other times the output is directly connected to ground.
+Voltage noise in current mode was measured with a 9.75 k$\Omega$ resistive load.
+$^b$Calculated as output range divided by resolution of output DAC (16 bit).}
+\end{table}
+
 \end{document}

### 2019-05-20 benchmarking/paper/draft2.Rnw
restructured library and stimjimPulser to c++ object called Stimjim, rather than just individual functions. this now roughly follows the arduino guidelines for library structure.

@@
 \documentclass[twocolumn]{article}
+%\documentclass{article}
 \usepackage[margin=2cm]{geometry}
@@
 Despite this, electronics that can provide precise current or voltage stimulation are expensive and closed-source.
-Here, we introduce stimjim, precise yet inexpensive (\$200) open-source hardware for electrical stimulation that combines both function generation and isolation.
+Here, we introduce Stimjim, a precise yet inexpensive (\$200) open-source instrument for electrical stimulation that combines both function generation and isolation.
 Stimjim provides microsecond temporal resolution with microampere or millivolt level precision on two isolated output channels.
@@
 Examples include Open Ephys\cite{siegle_open_2017}, Cyclops\cite{newman_optogenetic_2015}, OpenPCR\cite{noauthor_openpcr_nodate}, FlyPi\cite{chagas_100_2017}, Openspritzer\cite{forman_openspritzer:_2017}, and Audiomoth\cite{hill_audiomoth:_2018}.
-Many of these designs have been enabled by developments fueling the rapid growth of the maker community at large, including Arduino, Raspberry Pi, Teensy, openMV, pixyCam, open-source 3d printers like reprap.
+We now add Stimjim to this growing body of open hardware.
+Stimjim can replace an in vivo neural stimulator at a fraction of the cost, without compromising on quality.
+Furthermore, due to its entirely open design and software, Stimjim can be modified by end users to fit their specific needs.
 
@@
 Each stimulating channel includes a current source based on an improved Howland current pump, and a voltage source (an op-amp), driven by a 16-bit digital-to-analog converter (DAC).
-The final output of each channel is selected by a 4-way switch, such that either channel can be configured as a current output, voltage output, grounded, or disconnected (high-Z).
-To ensure the output is connected to a load and not either an open or short circuit, each channel also has a analog-to-digital converter able to read either the output voltage or the output current (via a sense resistor in series with the current output).
-Our design was made in Kicad, an open-source PCB CAD package.
+The final output of each channel is selected by a 4-way switch, such that either channel can be configured as a current output, voltage output, grounded, or disconnected.
+To ensure the stimulator is properly connected and not driving either an open or short circuit, each channel also has a analog-to-digital converter (ADC) able to read either the output voltage or the output current (via a sense resistor in series with the current output).
+Our circuit design was made in Kicad, an open-source printed circuit board (PCB) design program.
 Schematic, layout, bill of materials, and build instructions are included as supplemental materials and are also available in a git repository (https://bitbucket.org/natecermak/stimjim).
@@
 Stimjim's design is significantly less expensive than other commercial alternatives (Table 1).
-Its specifications are otherwise similar or better, with the notable exception that its compliance voltage is lower than other stimulators.
-This limits the load resistance that Stimjim can drive (for a given resistance R, Stimjim cannot output a current larger than $\frac{13.7V}{R}$).
-While this may be a problem for electrodes with very high impedances (e.g.
-sharp intracellular or patch electrodes), it is unlikely to be a problem for implanted stimulating electrodes.
+Its specifications are otherwise similar or better, with the exception that its compliance voltage is lower than other stimulators.
+This limits the load resistance that Stimjim can drive (for a given resistance $R$, Stimjim cannot output a current larger than $\frac{13.7V}{R}$).
+While this may be a problem for electrodes with very large impedances (for example, sharp intracellular or patch electrodes), it is unlikely to be a problem for implanted stimulating electrodes.
 
-\begin{table*}[t]
+Stimjim's software is written in C++ and is compatible with the Arduino programming environment.
+We provide both a library consisting of low-level functions (writing registers in the DACs or ADCs, or setting the channel mode), as well as a default program designed to output pulse trains.
+While end users are unlikely to need to adjust the library functions, they can use these functions to create new programs to run on the Stimjim - for example, generation of random pulse trains, or custom waveform outputs stored on the onboard SD card.
+
+\begin{table*}[h]
 \small{
@@
 Across the tested range of stimulation frequencies, Stimjim provided accurate and highly consistent inter-pulse intervals (IPIs; Fig.
-XA-C).
+\ref{fig:benchmarking}A-C).
 The standard deviation of interpulse intervals within a 1-second pulse train was always below one microsecond.
-Stimjim also produced pulse widths that were accurate and minimally variable (Fig XD-F).
+Stimjim also produced pulse widths that were accurate and minimally variable (Fig \ref{fig:benchmarking}D-F).
 Finally, we found Stimjim produced accurate voltage and current amplitudes.
 
-<<benchmarkFigure, echo=FALSE,fig.height=5.5,fig.width=5,fig.cap="\\label{fig:benchmarking} \\textbf{Electrical benchmarks show Stimjim provides microsecond temporal precision and millivolt- and microampere-scale amplitude precision.}", fig.align='center', out.width='1.1\\linewidth', fig.env="figure">>=
+<<benchmarkFigure, echo=FALSE,fig.height=8,fig.width=5,fig.cap="\\label{fig:benchmarking} \\textbf{Electrical benchmarks show Stimjim provides microsecond temporal precision and millivolt- and microampere-scale amplitude precision.}", fig.align='center', out.width='1\\linewidth', fig.env="figure">>=
 
@@
 
-layout(matrix(1:9,3,3,byrow=TRUE), widths=c(0.8,1,1))
-par(mar=c(4,1,2,1), mgp=c(2,0.3,0), cex.main=1.05, tck=-0.01)
+padj=-6
+layout(matrix(1:12,4,3,byrow=TRUE), widths=c(0.8,1,1))
+par(mar=c(4,1,2,1), mgp=c(2,0.3,0), cex.main=1.05, tck=-0.01,cex.axis=0.9)
 matplot(1e3/fs*1:nrow(ex1), dendRites::offset(ex1,15), col=1, type='l', lty=1, bty='n', yaxt='n', ylab='', xlab='time (ms)',
         main = "Varying frequency\n(Inter-pulse interval, IPI)"); 
-par(mar=c(4,4,2,1))
+mtext("A",side=2, padj=padj-0.5,las=2, line=0)
+par(mar=c(4,3,2,1))
 plot(paramMat[1:30,3]/1e3, sapply(res[seq(1,60,by=2)], function(x) x$IPI.summary[3])*1e3, yaxt='n',
@@
 abline(a=0,b=1)
+mtext("B",side=2, padj=padj,las=2)
 
@@
 axis(2,las=2)
+mtext("C",side=2, padj=padj,las=2)
 # 
@@
                 main = "Varying pulse width (PW)"); 
-par(mar=c(4,4,2,1))
+mtext("D",side=2, padj=padj-1,las=2)
+
+par(mar=c(4,3,2,1))
 plot(paramMat[31:59,7], sapply(res[seq(61,118,by=2)], function(x) x$PW.summary[3])*1e6/2,  yaxt='n',
@@
 axis(2,las=2)
+mtext("E",side=2, padj=padj,las=2)
 
@@
 axis(2,las=2)
+mtext("F",side=2, padj=padj,las=2)
 # plot(paramMat[31:59,7], sapply(res[seq(61,118,by=2)], function(x) diff(x$PW.summary[c(1,6)]))*1e6/2, 
@@
                 main = "Varying amplitude");
-par(mar=c(4,4,2,1))
+mtext("G",side=2, padj=padj,las=2)
+par(mar=c(4,3,2,1))
 x = c(paramMat[61:89,5], -paramMat[61:89,5])
@@
 axis(2,las=2)
+mtext("H",side=2, padj=padj,las=2)
 
@@
 axis(2,las=2)
-
+mtext("I",side=2, padj=padj,las=2)
 # plot(x, as.vector(y)-x, pch=16, xlab='', ylab='')
@@
 # axis(4, col='red')
+par(mar=c(4,2,2,1), mgp=c(1,0.3,0))
+dat = matrix(readBin(dataFiles[32], what='integer', n=1e7, size=2, endian = 'big'),ncol=2,byrow=TRUE)
+fp = min(which(dat[,1] > 100))
+ex4 = dat[fp + (-1e2:2e2),]*vpau
+matplot(1:nrow(ex4)/fs*1e6-50,ex4, main = 'Example short pulse', type='l', lty=1, xlab='Time (us)', ylab='Amplitude (V)', yaxt='n')
+axis(2,las=2)
+mtext("J",side=2, padj=padj,las=2, line=1)
 
+par(mar=c(4,3,2,1), mgp=c(2,0.3,0))
+dat = matrix(readBin(dataFiles[61], what='integer', n=1e7, size=2, endian = 'big'),ncol=2,byrow=TRUE)
+fp = min(which(dat[,1] > 10))
+ex5 = dat[fp + (-1e2:5e2),]*vpau
+matplot(1:nrow(ex5)/fs*1e6-50, ex5, main = 'Example small pulse', type='l', lty=1, xlab='Time (us)', ylab='Amplitude (V)', yaxt='n')
+axis(2,las=2)
+mtext("K",side=2, padj=padj,las=2)
+
+y = ex4[163:220,2]
+x=1:length(y)/2
+plot(x,y)
+mod = nls(y~a*(1-exp(-x/t))+b, start=list(a=-10,t=10,b=5))
+lines(x,predict(mod))
 @
@@
 Here, we trained two mice in a head-fixed variant of ICSS, in which animals could lick a sensor in order to obtain brain stimulation reward (BSR).
+We used a capacitive sensor attached to a small metal pole to detect licking, and every lick triggered a stimulus pulse train (0.5 seconds, initially 150 Hz and the minimal current at which animals would respond).
+To initially encourage licking, we placed a small amount of peanut butter on the metal sensor.
+After initial licking was reinforced by BSR, animals would continue licking long after the peanut butter was gone, including during the next session in which no peanut butter was offered.
 After animals had learned the licking behavior (usually within their first hour session), we varied the stimulation frequency and amplitude and assessed how it affected licking behavior.
@@
 For each animal, three frequency sweeps (highest to lowest frequency, 1 minute per frequency) were performed in a 45-minute session.
-Amplitude sweeps were performed in the same way.", cache=TRUE, fig.height=5, fig.width=5, fig.align='center'>>=
+Amplitude sweeps were performed in the same way.
+Each dot indicates a single 1-minute trial, and the solid black line shows the mean of all three trials at that frequency or amplitude.", cache=TRUE, fig.height=5, fig.width=5, fig.align='center'>>=
 

### 2019-05-20 lib/stimjim/src/Stimjim.cpp

@@ -1,21 +1,42 @@
 #include "Stimjim.h"
 
-// ------------- SPI setup ------------------------------------- //
-//AD5752 DAC runs at max 30MHz, SPI mode 1 or 2
-SPISettings SpiSettingsDAC(10000000, MSBFIRST, SPI_MODE1); // Does not seem to work at 20MHz!
-//AD7321 settings - clock starts high, data latch on falling edge
-SPISettings SpiSettingsADC(5000000, MSBFIRST, SPI_MODE2);  // Can't run SPI clock faster than 1MHz without errors
+void StimJim::begin(){
+
+  // ------------- SPI setup ------------------------------------- //
+  SPI.begin();
+
+  //AD5752 DAC runs at max 30MHz, SPI mode 1 or 2
+  SpiSettingsDAC = SPISettings(10000000, MSBFIRST, SPI_MODE1); 
+  //AD7321 settings - clock starts high, data latch on falling edge
+  SpiSettingsADC = SPISettings(5000000, MSBFIRST, SPI_MODE2);  
 
 
-// ------------- Compensation for various DC offsets ---------- //
-int currentOffsets[2];
-int voltageOffsets[2];
+  // ------------- Turn on LEDS ----------------------------------- //
+  pinMode(LED0, OUTPUT);
+  pinMode(LED1, OUTPUT);
+  digitalWrite(LED0, HIGH);
+  digitalWrite(LED1, HIGH);
 
-// --------- States of ADCs (set to read current or voltage?) ---//
-volatile bool adcSelectedInput[2];
+
+  // ------------- Setup output pins ------------------------------------ //
+  int pins[] = {NLDAC_0, NLDAC_1, CS0_0, CS1_0, CS0_1, CS1_1, OE0_0, OE1_0, OE0_1, OE1_1, LED0, LED1};
+  for (int i = 0; i < 12; i++) {
+    pinMode(pins[i], OUTPUT);
+    digitalWrite(pins[i], HIGH);
+  }
 
 
-void setOutputMode(byte channel, byte mode) {
+  setADCrange(10);
+  setupDACs();
+
+  getCurrentOffsets();
+  getVoltageOffsets();
+  delay(1000);
+  digitalWrite(LED0, LOW);
+  digitalWrite(LED1, LOW);
+}
+
+void StimJim::setOutputMode(byte channel, byte mode) {
[…]

### 2019-05-20 lib/stimjim/src/Stimjim.h

@@ -32,28 +32,32 @@
 #define LED1 20
 #define MISO_0 8
 #define MISO_1 12
+ 
 
 
 
+class StimJim {
 
+  public:
+    void begin();
+    void setOutputMode(byte channel, byte mode);
+    void setADCrange(float range);
+    void writeToDacs(int16_t amp0, int16_t amp1);
+    void writeToDac(byte channel, int16_t amp);
+    int readADC(byte channel, byte line);
+    void getCurrentOffsets();
+    void getVoltageOffsets();
 
-// ------------ Function prototypes --------------------------- //
-void setOutputMode(byte channel, byte mode);
-void setADCrange(float range);
-void setupDACs();
-void writeToDacs(int16_t amp0, int16_t amp1);
-void writeToDac(byte channel, int16_t amp);
-int readADC(byte channel, byte line);
-void getCurrentOffsets();
-void getVoltageOffsets();
+    int currentOffsets[2]; //  Compensation for various DC offsets
+    int voltageOffsets[2];
 
+  private:
+    void setupDACs();
+    volatile bool adcSelectedInput[2]; // --------- States of ADCs (set to read current or voltage?) ---//
+    SPISettings SpiSettingsDAC, SpiSettingsADC;
 
-// ------------- Compensation for various DC offsets ---------- //
-extern int currentOffsets[2];
-extern int voltageOffsets[2];
+};
 
-// --------- States of ADCs (set to read current or voltage?) ---//
-extern volatile bool adcSelectedInput[2];
-
+extern StimJim Stimjim;
 
 #endif /* ! defined __STIMJIM_H__ */

### 2019-05-28 benchmarking/paper/draft2.Rnw
realized my 5k resistors were actually 5.05k. modified gain multiplier in library and schematic (kicad and pdf) to say 5.05k.

@@
 
-projectDirectory = '/home/x/Desktop/stimjim/benchmarking/'
+projectDirectory = '/home/x/Dropbox (Technion Dropbox)/stimjim/benchmarking/'
 setwd(projectDirectory)
@@
 
-    sh.median = median(x[which(isStageHigh>0)])*vpau
-    sh.sd = sd(x[which(mmand::erode(isStageHigh, rep(1,13))>0)])*vpau
-    sl.median = median(x[which(isStageLow>0)])*vpau
-    sl.sd = sd(x[which(mmand::erode(isStageLow, rep(1,13))>0)])*vpau
+    ishe = which(mmand::erode(isStageHigh>0, rep(1,31)) > 0)
+    isle = which(mmand::erode(isStageLow>0, rep(1,31)) > 0 )
+    sh.median = median(x[ishe])*vpau
+    sh.sd = sd(x[ishe])*vpau
+    sl.median = median(x[isle])*vpau
+    sl.sd = sd(x[isle])*vpau
 
@@
 
-y = ex4[163:220,2]
-x=1:length(y)/2
-plot(x,y)
-mod = nls(y~a*(1-exp(-x/t))+b, start=list(a=-10,t=10,b=5))
-lines(x,predict(mod))
+#y = ex4[163:220,2]
+#x=1:length(y)/2
+#plot(x,y)
+#mod = nls(y~a*(1-exp(-x/t))+b, start=list(a=-10,t=10,b=5))
+#lines(x,predict(mod))
 @

### 2019-05-28 lib/stimjim/src/Stimjim.h

@@ -13,7 +13,7 @@
 
 #define MICROAMPS_PER_DAC 0.1017            // 20V * (1/(3000 V/A)) / 2^16 = 0.1uA / DAC unit
 #define MICROAMPS_PER_ADC 0.85              //  (1/(100*(1+49.9k/1.8k) V/A)) * 20V / 2^13 = 0.85
-#define MILLIVOLTS_PER_DAC 0.458             // 20V / 2^16 * gain of 1.5 ~= 0.44 mV / DAC unit
+#define MILLIVOLTS_PER_DAC 0.4593            // 20V / 2^16 * gain of 1.505 ~= 0.44 mV / DAC unit
 #define MILLIVOLTS_PER_ADC 2.44             // 20V / 2^13 = 2.44 mV / ADC unit
 
 #define OE0_0 6

### 2019-05-28 benchmarking/paper/draft2.Rnw
updates to analysis - pulse time window for calculating amplitude updated.

@@
 
-    ishe = which(mmand::erode(isStageHigh>0, rep(1,31)) > 0)
-    isle = which(mmand::erode(isStageLow>0, rep(1,31)) > 0 )
-    sh.median = median(x[ishe])*vpau
+    ishe = which(mmand::erode(mmand::erode(isStageHigh>0, c(rep(1,61),rep(0,60))), rep(1,61)) > 0)
+    isle = which(mmand::erode(mmand::erode(isStageLow>0, c(rep(1,61),rep(0,60))), rep(1,61)) > 0)
+    sh.median = mean(x[ishe])*vpau
     sh.sd = sd(x[ishe])*vpau
-    sl.median = median(x[isle])*vpau
+    sl.median = mean(x[isle])*vpau
     sl.sd = sd(x[isle])*vpau
@@
 \section*{Abstract}
-Electrical stimulation is a powerful tool to perturb and understand the function of neurons and neural circuits.
-Despite this, electronics that can provide precise current or voltage stimulation are expensive and closed-source.
-Here, we introduce Stimjim, a precise yet inexpensive (\$200) open-source instrument for electrical stimulation that combines both function generation and isolation.
-Stimjim provides microsecond temporal resolution with microampere or millivolt level precision on two isolated output channels.
-To show the utility of Stimjim, we demonstrate training of mice to perform intracranial self-stimulation (ICSS) for brain stimulation reward (BSR) in a novel head-fixed paradigm, and that interest in the reward can be tuned by adjusting either stimulation frequency or amplitude.
+Electrical stimulation is a simple and powerful tool to perturb and understand the function of neurons and neural circuits.
+Despite this, devices that can provide precise current or voltage stimulation are expensive and closed-source.
+Here, we introduce Stimjim, a capable and inexpensive (\$200) open-source instrument for electrical stimulation that combines both function generation and electrical isolation.
+Stimjim provides microsecond temporal resolution with microampere or millivolt level precision on two electrically isolated output channels.
+To show the utility of Stimjim, we use it to train mice to perform intracranial self-stimulation (ICSS) for brain stimulation reward (BSR) in a novel head-fixed paradigm.
+We show that reward-seeking behavior can be smoothly tuned by adjusting either stimulation frequency or amplitude.
 
 \section*{Introduction}
-Electrical stimulation of neural tissue is an invaluable research tool.
+Electrical stimulation of neural tissue is an invaluable and ubiquitous research tool.
 Over the past 150 years, it has enabled an understanding of the function of various brain regions by directly causing neurons in those regions to fire \cite{hitzig_uber_1870,ferrier_functions_1876,penfield_somatic_1937}.
-More recently, it has also found important clinicial applications in neurological disorders including Parkinson's disease\cite{deuschl_randomized_2006} and depression\cite{mayberg_deep_2005}.
-However, to date, the hardware enabling precise control over current and voltage stimulation remains generally expensive and closed source.
+More recently, it has also found important clinical applications in neurological disorders including Parkinson's disease\cite{deuschl_randomized_2006} and depression\cite{mayberg_deep_2005}.
+However, to date, the hardware enabling precise control over current and voltage stimulation generally remains expensive and closed source.
 
@@
 We now add Stimjim to this growing body of open hardware.
-Stimjim can replace an in vivo neural stimulator at a fraction of the cost, without compromising on quality.
+Stimjim replaces \textit{in vivo} neural stimulators at a fraction of the cost, without compromising on quality.
 Furthermore, due to its entirely open design and software, Stimjim can be modified by end users to fit their specific needs.
@@
 \section*{Results}
-\textbf{Design.} We developed Stimjim to be an precision stimulus generator and isolator in a single package.
+\textbf{Design.} We developed Stimjim to be a precise, isolated stimulus generator.
 Stimjim is based on the Teensy 3.5 microcontroller, which utilizes a 32-bit Arm Cortex-M4 processor running at 120MHz.
@@
 The final output of each channel is selected by a 4-way switch, such that either channel can be configured as a current output, voltage output, grounded, or disconnected.
-To ensure the stimulator is properly connected and not driving either an open or short circuit, each channel also has a analog-to-digital converter (ADC) able to read either the output voltage or the output current (via a sense resistor in series with the current output).
+To ensure the stimulator is properly connected (a common issue with experiments in freely moving animals) and to verify required stimulus current or voltage amplitudes, each channel also has a analog-to-digital converter (ADC) able to read either the output voltage or the output current (via a low-value sense resistor in series with the current output).
 Our circuit design was made in Kicad, an open-source printed circuit board (PCB) design program.
@@
 x = c(paramMat[61:89,5], -paramMat[61:89,5])
-y = t(sapply(res[seq(121,178,by=2)], function(x) c(x$sh.median,x$sl.median))*1e3)
+y = t(sapply(res[seq(121,178,by=2)], function(x) c(x$sh.median,x$sl.median))*1e3) * 0.458/0.4593
 plot(x/1000,y-x, main='Voltage accuracy', xlab='Target amplitude (V)', ylab='Measured error (mV)', cex=0.75, yaxt='n', ylim=c(-30,30))
@@
 \ref{fig:icss}).
+
+Notably, we observed clear differences between the two animals studied here.
+Mouse 1 shows a rather linear response to either increasing frequency or increasing amplitude, whereas mouse 2 had a more "digital" response akin to passing an activation threshold.
+Notably, maximal licking rates were comparable between the two animals.
+Such differences are likely due to electrode placement, although they may also reflect intrinsically different personalities between the two animals.
+Stimjim provides a precise and cost-effective means to scan the space of stimulation patterns, which could be useful to ensure all animals are given stimuli yielding the same response level.
 

### 2019-06-05 lib/stimjim/src/Stimjim.h
switched output channels in stimjim library to be consistent with front panel. (TODO: schematic still inconsistent). Reorganized BOM.

@@ -16,22 +16,22 @@
 #define MILLIVOLTS_PER_DAC 0.4593            // 20V / 2^16 * gain of 1.505 ~= 0.44 mV / DAC unit
 #define MILLIVOLTS_PER_ADC 2.44             // 20V / 2^13 = 2.44 mV / ADC unit
 
-#define OE0_0 6
-#define OE1_0 5
-#define OE0_1 1
-#define OE1_1 0
-#define CS0_0 9
-#define CS1_0 7
-#define CS0_1 3
-#define CS1_1 2
-#define NLDAC_1 4
-#define NLDAC_0 10
+#define OE0_1 6
+#define OE1_1 5
+#define OE0_0 1
+#define OE1_0 0
+#define CS0_1 9
+#define CS1_1 7
+#define CS0_0 3
+#define CS1_0 2
+#define NLDAC_0 4
+#define NLDAC_1 10
 #define IN0 22
 #define IN1 23
 #define LED0 21
 #define LED1 20
-#define MISO_0 8
-#define MISO_1 12
+#define MISO_1 8
+#define MISO_0 12
  
 
 

### 2019-06-05 benchmarking/paper/draft2.Rnw
added figure captions, more references to analysis.

@@
 \usepackage[super]{natbib}
+\usepackage[font={small}]{caption}
 \begin{document}
@@
     isle = which(mmand::erode(mmand::erode(isStageLow>0, c(rep(1,61),rep(0,60))), rep(1,61)) > 0)
-    sh.median = mean(x[ishe])*vpau
+    sh.median = median(x[ishe])*vpau
+    sh.mean = mean(x[ishe])*vpau
     sh.sd = sd(x[ishe])*vpau
-    sl.median = mean(x[isle])*vpau
+    sl.median = median(x[isle])*vpau
+    sl.mean = mean(x[isle])*vpau
     sl.sd = sd(x[isle])*vpau
@@
              PW.sd=PW.sd, sh.median=sh.median, sh.sd=sh.sd, 
-             sl.median=sl.median, sl.sd=sl.sd, pulseMat=pulseMat)
+             sl.median=sl.median, sl.sd=sl.sd, sl.mean=sl.mean, sh.mean=sh.mean, pulseMat=pulseMat)
   })
@@
 To show the utility of Stimjim, we use it to train mice to perform intracranial self-stimulation (ICSS) for brain stimulation reward (BSR) in a novel head-fixed paradigm.
-We show that reward-seeking behavior can be smoothly tuned by adjusting either stimulation frequency or amplitude.
+By varying either the output frequency or amplitude, Stimjim enables the experimenter to smoothly tune the strength of reward-seeking behavior.
 
@@
 
-In contrast, there has been a recent push within the scientific community to produce open labware - open source hardware and software replacements for a variety of common laboratory tasks.
-Examples include Open Ephys\cite{siegle_open_2017}, Cyclops\cite{newman_optogenetic_2015}, OpenPCR\cite{noauthor_openpcr_nodate}, FlyPi\cite{chagas_100_2017}, Openspritzer\cite{forman_openspritzer:_2017}, and Audiomoth\cite{hill_audiomoth:_2018}.
+In contrast, there has been a recent push within the scientific community to produce open labware - open source hardware and software replacements for a variety of common laboratory tasks\cite{pearce_open-source_2013,baden_open_2015} Among many others, examples include software and hardware for:
+\begin{itemize}
+\item recording or stimulating neurons (e.g.
+Open Ephys\cite{siegle_open_2017,buccino_open_2018}, Cyclops\cite{newman_optogenetic_2015}, Neurorighter\cite{newman_closed-loop_2013}, Bonsai\cite{lopes_bonsai:_2015})
+\item amplifying DNA (e.g., OpenPCR\cite{noauthor_openpcr_nodate}) or preparing single cells for DNA sequencing \cite{booeshaghi_design_2019}
+\item imaging microscopic samples (e.g., FlyPi\cite{chagas_100_2017})
+\item quantifying chemical reactions or microbial growth in 96-well plates (e.g., \cite{szymula_open-source_2019})
+\item growing and evolving bacteria (e.g. \cite{takahashi_low_2015}) 
+\item ecological monitoring (e.g., Audiomoth\cite{hill_audiomoth:_2018})
+\end{itemize}
 We now add Stimjim to this growing body of open hardware.
@@
 
-<<benchmarkFigure, echo=FALSE,fig.height=8,fig.width=5,fig.cap="\\label{fig:benchmarking} \\textbf{Electrical benchmarks show Stimjim provides microsecond temporal precision and millivolt- and microampere-scale amplitude precision.}", fig.align='center', out.width='1\\linewidth', fig.env="figure">>=
+<<benchmarkFigure, echo=FALSE,fig.height=8,fig.width=5,fig.cap="\\label{fig:benchmarking} \\textbf{Electrical benchmarks show Stimjim provides microsecond temporal precision and millivolt- and microampere-scale amplitude precision.} \\\\ (A) Example pulse trains with varying inter-pulse interval (IPI).
+\\\\ (B) Median IPI in a 1-second pulse train measured on a high-speed data acquisition device, plotted against target IPI.
+Stimjim provides precise frequency control.
+\\\\ (C) Standard deviation of inter-pulse intervals measured in a 1-second pulse train.
+\\\\ (D) Example pulse trains with varying pulse width (PW).
+\\\\ (E) Median measured PW vs target PW, showing Stimjim has precise pulse width control.
+\\\\ (F) Standard deviation of pulse widths over 100 pulses.
+\\\\ (G) Example pulses with varying amplitude.\\\\ (H) Error in amplitude of voltage pulses (100 us, 1 kHz) vs target amplitude.\\\\ (I) same as (H) but for current pulses.
+\\\\ (J) Example of a fast biphasic (30 us/phase) pulse.
+Red line is a 500uA current pulse (with a 9.75k resistive load), black line is a 5V pulse.
+\\\\ (K) Example of a small-amplitude pulse.
+Red line is a 20uA current pulse (with a 9.75k resistive load), black line is a 0.2V pulse.", fig.align='center', out.width='1\\linewidth', fig.env="figure">>=
 
@@
 mtext("C",side=2, padj=padj,las=2)
-# 
-# plot(paramMat[1:30,3], sapply(res[seq(1,60,by=2)], function(x) diff(x$IPI.summary[c(1,6)]))*1e6, 
-#   log='x', main='IPI range', cex=0.75, xlab='Target IPI', ylab='Range of measured IPIs')
-# points(paramMat[1:30,3], sapply(res[seq(2,60,by=2)], function(x) diff(x$IPI.summary[c(1,6)]))*1e6,
-#   col='red', cex=0.75)
 
@@
 mtext("F",side=2, padj=padj,las=2)
-# plot(paramMat[31:59,7], sapply(res[seq(61,118,by=2)], function(x) diff(x$PW.summary[c(1,6)]))*1e6/2, 
-#   log='x', main='PW range', ylab='PW range', cex=0.75)
-# points(paramMat[31:59,7], sapply(res[seq(62,118,by=2)], function(x) diff(x$PW.summary[c(1,6)]))*1e6/2,
-#   col='red', cex=0.75)
 
@@
 mtext("G",side=2, padj=padj,las=2)
+
 par(mar=c(4,3,2,1))
 x = c(paramMat[61:89,5], -paramMat[61:89,5])
-y = t(sapply(res[seq(121,178,by=2)], function(x) c(x$sh.median,x$sl.median))*1e3) * 0.458/0.4593
+y = t(sapply(res[seq(121,178,by=2)], function(x) c(x$sh.mean,x$sl.mean))*1e3)
 plot(x/1000,y-x, main='Voltage accuracy', xlab='Target amplitude (V)', ylab='Measured error (mV)', cex=0.75, yaxt='n', ylim=c(-30,30))
@@
 x2 = c(paramMat[61:90,6], -paramMat[61:90,6])
-y2 = t(sapply(res[seq(122,180,by=2)], function(x) c(x$sh.median,x$sl.median))/9750*1e6)
+y2 = t(sapply(res[seq(122,180,by=2)], function(x) c(x$sh.mean,x$sl.mean))/9750*1e6)
 plot(x2/1000,y2-x2, main='Current accuracy', xlab='Target amplitude (mA)', ylab='Measured error (uA)', cex=0.75, col='red', yaxt='n',ylim=c(-5,5))
@@
 mtext("I",side=2, padj=padj,las=2)
-# plot(x, as.vector(y)-x, pch=16, xlab='', ylab='')
-# par(new=TRUE)
-# plot(x2, as.vector(y2)-x2, pch=16, axes=F, col='red', xlab='',ylab='')
-# axis(1, col='red',line=2); 
-# axis(4, col='red')
+
 par(mar=c(4,2,2,1), mgp=c(1,0.3,0))
@@
 
-#y = ex4[163:220,2]
-#x=1:length(y)/2
-#plot(x,y)
-#mod = nls(y~a*(1-exp(-x/t))+b, start=list(a=-10,t=10,b=5))
-#lines(x,predict(mod))
 @
@@
 
-\bibliographystyle{nat}
+\bibliographystyle{unsrt}
 \bibliography{stimjimBibliography}
@@
  Current output step size (current mode)$^b$ & 0.1         & uA        \\ 
+ Output impedance (voltage mode)$^c$         & 180         & $\Omega$ \\ 
  Trigger latency                             & 23          & us        \\
@@
 Voltage noise in current mode was measured with a 9.75 k$\Omega$ resistive load.
-$^b$Calculated as output range divided by resolution of output DAC (16 bit).}
+$^b$Calculated as output range divided by resolution of output DAC (16 bit).
+$^c$ value from from datasheet for Vishay DG509B (output switch).
+}
 \end{table}

### 2019-06-10 benchmarking/paper/draft2.Rnw
more updates to draft

@@
 \usepackage[font={small}]{caption}
+
+\newenvironment{itemize*}%
+  {\begin{itemize}%
+    \setlength{\itemsep}{0pt}%
+    \setlength{\parskip}{0pt}}%
+  {\end{itemize}}
+
 \begin{document}
@@
     pulseMat = sapply(pulseStartIndices, function(j) x[j+(-100:1000)])
-    list(IPI.summary=IPI.summary, IPI.sd=IPI.sd, PW.summary=PW.summary, 
+    list(IPI.summary=IPI.summary, IPI.sd=IPI.sd, PW.summary=PW.summary, PW=PW, IPI=IPI,
              PW.sd=PW.sd, sh.median=sh.median, sh.sd=sh.sd, 
@@
 
-In contrast, there has been a recent push within the scientific community to produce open labware - open source hardware and software replacements for a variety of common laboratory tasks\cite{pearce_open-source_2013,baden_open_2015} Among many others, examples include software and hardware for:
-\begin{itemize}
+In contrast, there has been a recent push within the scientific community to produce open labware - open source hardware and software replacements for a variety of common laboratory tasks\cite{pearce_open-source_2013,baden_open_2015}.
+Among many others, examples include software and hardware for:
+\begin{itemize*}
 \item recording or stimulating neurons (e.g.
@@
 \item imaging microscopic samples (e.g., FlyPi\cite{chagas_100_2017})
-\item quantifying chemical reactions or microbial growth in 96-well plates (e.g., \cite{szymula_open-source_2019})
-\item growing and evolving bacteria (e.g. \cite{takahashi_low_2015}) 
+\item quantifying chemical reactions or microbial growth in 96-well plates\cite{szymula_open-source_2019}
+\item growing and evolving bacteria\cite{takahashi_low_2015} 
 \item ecological monitoring (e.g., Audiomoth\cite{hill_audiomoth:_2018})
-\end{itemize}
+\end{itemize*}
 We now add Stimjim to this growing body of open hardware.
@@
 
-Stimjim's design is significantly less expensive than other commercial alternatives (Table 1).
-Its specifications are otherwise similar or better, with the exception that its compliance voltage is lower than other stimulators.
+Stimjim's design compares well against alternatives (Table 1).
+It is order-of-magnitude less expensive than most commercial alternatives.
+In comparison with commercial stimulators, its only drawback is that its compliance voltage is lower.
 This limits the load resistance that Stimjim can drive (for a given resistance $R$, Stimjim cannot output a current larger than $\frac{13.7V}{R}$).
-While this may be a problem for electrodes with very large impedances (for example, sharp intracellular or patch electrodes), it is unlikely to be a problem for implanted stimulating electrodes.
+Stimjim's cost is most similar to PulsePal 2, an open-source programmable pulse generator \cite{sanders_low-cost_2014}.
+However, it expands on PulsePal's capabilities by including electrical isolation, current output mode, and on-board monitoring of output currents/voltages.
+Further electrical characteristics of Stimjim are given in Table 2.
 
-Stimjim's software is written in C++ and is compatible with the Arduino programming environment.
-We provide both a library consisting of low-level functions (writing registers in the DACs or ADCs, or setting the channel mode), as well as a default program designed to output pulse trains.
-While end users are unlikely to need to adjust the library functions, they can use these functions to create new programs to run on the Stimjim - for example, generation of random pulse trains, or custom waveform outputs stored on the onboard SD card.
+Stimjim's software is written in the Arduino language (C++) and is compatible with the Arduino programming environment.
+We provide an Arduino library consisting of low-level functions (writing registers in the DACs or ADCs, or setting the channel mode).
+Library functions enable users to create new programs to run on the Stimjim - for example, generation of custom waveform outputs stored on the onboard SD card.
+We also provide a default program designed to output pulse trains.
+As with previous pulse generators\cite{sanders_low-cost_2014}, pulse trains are parameterized based on their output mode (current or voltage), frequency, duration, and the amplitude of each phase of the pulse itself.
 
@@
 
-Stimjim proved exceptionally accurate, providing microsecond temporal resolution and millivolt- and microampere- amplitude resolution.
+Stimjim proved exceedingly accurate, providing microsecond temporal resolution and millivolt- and microampere- amplitude resolution.
 Across the tested range of stimulation frequencies, Stimjim provided accurate and highly consistent inter-pulse intervals (IPIs; Fig.
@@
 Stimjim also produced pulse widths that were accurate and minimally variable (Fig \ref{fig:benchmarking}D-F).
-Finally, we found Stimjim produced accurate voltage and current amplitudes.
+We then assessed pulse amplitudes across a range of settings to ensure negligible DC offsets and proper gains.
+From -10V to +10V (the range of our test equipment), Stimjim produced accurate voltage and current amplitudes, with maximal errors of less than 10mV and 3uA (Fig \ref{fig:benchmarking}G-I).
 
-<<benchmarkFigure, echo=FALSE,fig.height=8,fig.width=5,fig.cap="\\label{fig:benchmarking} \\textbf{Electrical benchmarks show Stimjim provides microsecond temporal precision and millivolt- and microampere-scale amplitude precision.} \\\\ (A) Example pulse trains with varying inter-pulse interval (IPI).
-\\\\ (B) Median IPI in a 1-second pulse train measured on a high-speed data acquisition device, plotted against target IPI.
-Stimjim provides precise frequency control.
-\\\\ (C) Standard deviation of inter-pulse intervals measured in a 1-second pulse train.
-\\\\ (D) Example pulse trains with varying pulse width (PW).
-\\\\ (E) Median measured PW vs target PW, showing Stimjim has precise pulse width control.
-\\\\ (F) Standard deviation of pulse widths over 100 pulses.
-\\\\ (G) Example pulses with varying amplitude.\\\\ (H) Error in amplitude of voltage pulses (100 us, 1 kHz) vs target amplitude.\\\\ (I) same as (H) but for current pulses.
-\\\\ (J) Example of a fast biphasic (30 us/phase) pulse.
+<<benchmarkFigure, echo=FALSE,fig.height=8,fig.width=5,fig.cap="\\label{fig:benchmarking} \\textbf{Electrical benchmarks show Stimjim provides microsecond temporal precision and millivolt- and microampere-scale amplitude precision.}  (A) Example pulse trains with varying inter-pulse interval (IPI).
+(B) Median IPI in a 1-second pulse train measured on a high-speed data acquisition device, plotted against target IPI.
+Red dots show current output, black dots show voltage output.
+Solid line shows equality.
+(C) Boxplot of IPI errors measured in a 1-second pulse train.
+Note that the due to the 2MHz sampling rate of our test equipment, the box representing the interquartile range often has zero height.
+Whiskers are drawn to most extreme values, irrespective of interquartile range.
+(D) Example pulse trains with varying pulse width (PW).
+(E) Median measured PW vs target PW.
+Red dots show current output, black dots show voltage output.
+Solid line shows equality.
+(F) Boxplots of pulse width errors over 100 pulses.
+Conventions are the same as for (C).
+(G) Example pulses with varying amplitude.
+(H) Error in amplitude of voltage pulses (100 us, 1 kHz) vs target amplitude.
+(I) same as (H) but for current pulses.
+(J) Example of a fast biphasic (30 us/phase) pulse.
 Red line is a 500uA current pulse (with a 9.75k resistive load), black line is a 5V pulse.
-\\\\ (K) Example of a small-amplitude pulse.
+(K) Example of a small-amplitude pulse.
 Red line is a 20uA current pulse (with a 9.75k resistive load), black line is a 0.2V pulse.", fig.align='center', out.width='1\\linewidth', fig.env="figure">>=
@@
         main = "Varying frequency\n(Inter-pulse interval, IPI)"); 
-mtext("A",side=2, padj=padj-0.5,las=2, line=0)
+mtext("A",side=2, padj=padj-0.2,las=2, line=0)
 par(mar=c(4,3,2,1))
@@
 
-plot(paramMat[1:30,3]/1e3, sapply(res[seq(1,60,by=2)], function(x) x$IPI.sd)*1e6, yaxt='n',
-  log='x', main='', cex=0.75, xlab='Target IPI (ms)', ylab='IPI standard deviation (us)')
-points(paramMat[1:30,3]/1e3, sapply(res[seq(2,60,by=2)], function(x) x$IPI.sd)*1e6,
-  col='red', cex=0.75)
+boxplot(lapply(seq(1,60), function(i) res[[i]]$IPI*1e6 - paramMat[(i+1)/2,3]), axes=FALSE, range=0, lty=1, staplewex=0,
+  log='', main='', cex=0.75, xlab='Target IPI (ms)', ylab='IPI error (us)', col=1:2, border=1:2, ylim=c(-4,4))
 axis(2,las=2)
+axis(1,las=2, at=seq(1.5,60,by=2), labels=signif(paramMat[1:30,3]/1e3, 3))
 mtext("C",side=2, padj=padj,las=2)
@@
 
-plot(paramMat[31:59,7], sapply(res[seq(61,118,by=2)], function(x) x$PW.sd)*1e6/2, yaxt='n',
-  log='x', main='', xlab='Target PW (us)', ylab='PW standard deviation (us)', cex=0.75)
-points(paramMat[31:59,7], sapply(res[seq(62,118,by=2)], function(x) x$PW.sd)*1e6/2,
-  col='red', cex=0.75)
+boxplot(lapply(seq(61,118), function(i) res[[i]]$PW*1e6/2 - paramMat[(i+1)/2,7]), axes=FALSE, range=0, lty=1, staplewex=0,
+  log='', main='', cex=0.75, xlab='Target PW (ms)', ylab='PW error (us)', col=1:2, border=1:2)
 axis(2,las=2)
+axis(1,las=2, at=seq(1.5,60,by=2), labels=signif(paramMat[1:30,3]/1e3, 3))
 mtext("F",side=2, padj=padj,las=2)
@@
 ex4 = dat[fp + (-1e2:2e2),]*vpau
-matplot(1:nrow(ex4)/fs*1e6-50,ex4, main = 'Example short pulse', type='l', lty=1, xlab='Time (us)', ylab='Amplitude (V)', yaxt='n')
+matplot(1:nrow(ex4)/fs*1e6-50,ex4, main = 'Example fast pulse', type='l', lty=1, xlab='Time (us)', ylab='Amplitude (V)', yaxt='n')
 axis(2,las=2)
@@
 To initially encourage licking, we placed a small amount of peanut butter on the metal sensor.
-After initial licking was reinforced by BSR, animals would continue licking long after the peanut butter was gone, including during the next session in which no peanut butter was offered.
+After ini tial licking was reinforced by BSR, animals would continue licking long after the peanut butter was gone, including during the next session in which no peanut butter was offered.
 After animals had learned the licking behavior (usually within their first hour session), we varied the stimulation frequency and amplitude and assessed how it affected licking behavior.
@@
 
-\section*{Conclusion}
-
-\section*{Acknowledgments}
-We thank the Schiller lab for discussions and help with preliminary testing. This work was supported in part by the Zuckerman STEM Leadership Program (NC).
-
-\section*{Methods}
-\textbf{Electrode implantation}.
-Monopolar electrodes (Plastics1, MS303/2-AIU/SPC) were implanted above the medial forebrain bundle according to the protocol in reference \cite{carlezon_intracranial_2007}.
-Additionally, a 3D-printed headpost was affixed to the animal's skull by dental cement to enable head fixation.
-All animal procedures were in accordance with guidelines established bythe NIH on the care and use of animals in research and were confirmed by the Technion Institutional Animal Care and Use Committee (IL-012-01-18, valid until 10/4/2022).
-
-\section*{Supplemental materials}
-BOM, gerbers, schematic, link to bitbucket
-
-\bibliographystyle{unsrt}
-\bibliography{stimjimBibliography}
-
 \begin{table}
@@
 \hline
- Slew rate (voltage mode)                    & 4.1         & V/us     \\
- Slew rate (current mode)                    & 0.8         & V/us     \\
+ Slew rate (voltage mode)                    & 4.1         & V/$\mu$s     \\
+ Slew rate (current mode)                    & 0.8         & V/$\mu$s     \\
  Output voltage noise (voltage mode)$^a$     & 1           & mV rms    \\
@@
                                              & 3           & mVpp      \\
- Voltage output step size (voltage mode)$^b$ & 0.45        & V         \\
- Current output step size (current mode)$^b$ & 0.1         & uA        \\ 
+ Min. voltage increment$^b$                  & 0.45        & V         \\
+ Min. current increment$^b$                  & 0.1         & $\mu$A        \\ 
  Output impedance (voltage mode)$^c$         & 180         & $\Omega$ \\ 
- Trigger latency                             & 23          & us        \\
+ Trigger latency                             & 23          & $\mu$s        \\
  \hline
@@
 
+\section*{Conclusion}
+We have introduced Stimjim, an inexpensive yet precise open-source stimulator for neuroscience.
+At a cost of roughly \$200 USD for parts, Stimjim is order of magnitude less expensive than commercial, proprietary alternatives.
+It offers microsecond temporal control of current and voltage with millivolt/microampere precision.
+
+Stimjim's low cost opens up many potential applications, such as learning paradigms that involve direct electrical stimulation.
+Stimjim's open source nature makes it straightforward for researchers to customize the stimulation parameters and use Stimjim in closed loop experiments.
+Furthermore, researchers who were previously limited to training only one animal at a time due to possessing only a single stimulator could now train 10 animals simultaneously for comparable cost.
+
+\section*{Acknowledgments}
+We thank the Schiller lab for discussions and help with preliminary testing. This work was supported in part by the Zuckerman STEM Leadership Program (NC).
+
+\section*{Methods}
+\textbf{Electrode implantation}.
+Monopolar electrodes (Plastics1, MS303/2-AIU/SPC, coated stainless steel, 200um diameter) were implanted above the medial forebrain bundle according to the protocol in reference \cite{carlezon_intracranial_2007}.
+The ground was implanted in the contralateral cortex.
+Additionally, a 3D-printed headpost was affixed to the animal's skull by dental cement to enable head fixation.
+Typical resistance between connector pins after implantation was 20-30k$\Omega$.
+All animal procedures were in accordance with guidelines established by the NIH on the care and use of animals in research and were confirmed by the Technion Institutional Animal Care and Use Committee (IL-012-01-18, valid until 10/4/2022).
+
+\section*{Supplemental materials}
+In case of further modifications, the most up-to-date details regarding Stimjim will be available at https://bitbucket.org/natecermak/stimjim.
+The following are included as supplemental materials for the version of Stimjim documented here (v0.17).
+\begin{itemize*}
+\item Bill of materials: stimjim\_bom.xlsx 
+\item Fabrication files: gerbers.zip 
+\item Schematic: schematic.pdf 
+\end{itemize*}
+
+\newpage
+
+\bibliographystyle{unsrt}
+\bibliography{stimjimBibliography}
+
 \end{document}

### 2019-06-17 benchmarking/paper/draft2.Rnw
minor fixes in draft, mainly to units

@@
 In contrast, there has been a recent push within the scientific community to produce open labware - open source hardware and software replacements for a variety of common laboratory tasks\cite{pearce_open-source_2013,baden_open_2015}.
-Among many others, examples include software and hardware for:
+Examples in life sciences include software and hardware for:
 \begin{itemize*}
@@
 We now add Stimjim to this growing body of open hardware.
-Stimjim replaces \textit{in vivo} neural stimulators at a fraction of the cost, without compromising on quality.
+Stimjim replaces commercial \textit{in vivo} neural stimulators at a fraction of the cost, without compromising on quality.
 Furthermore, due to its entirely open design and software, Stimjim can be modified by end users to fit their specific needs.
@@
 Stimjim is based on the Teensy 3.5 microcontroller, which utilizes a 32-bit Arm Cortex-M4 processor running at 120MHz.
-Each stimulating channel includes a current source based on an improved Howland current pump, and a voltage source (an op-amp), driven by a 16-bit digital-to-analog converter (DAC).
+Each stimulating channel includes a current source based on an improved Howland current pump\cite{texas_instruments_-1515_2008}, and a voltage source (an op-amp), driven by a 16-bit digital-to-analog converter (DAC).
 The final output of each channel is selected by a 4-way switch, such that either channel can be configured as a current output, voltage output, grounded, or disconnected.
@@
 In comparison with commercial stimulators, its only drawback is that its compliance voltage is lower.
-This limits the load resistance that Stimjim can drive (for a given resistance $R$, Stimjim cannot output a current larger than $\frac{13.7V}{R}$).
+This limits the load resistance that Stimjim can drive - for a given resistance $R$, Stimjim cannot output a current larger than $\frac{13.7~V}{R}$.
 Stimjim's cost is most similar to PulsePal 2, an open-source programmable pulse generator \cite{sanders_low-cost_2014}.
@@
 We also provide a default program designed to output pulse trains.
-As with previous pulse generators\cite{sanders_low-cost_2014}, pulse trains are parameterized based on their output mode (current or voltage), frequency, duration, and the amplitude of each phase of the pulse itself.
+Pulse trains are parameterized based on their output mode (current or voltage), frequency, duration, and the amplitude of each phase of the pulse itself.
 
@@
  Onboard measurement            & Y             & N                &  N                                & Optional                 & Y*             \\
- Compliance voltage             & $\pm$ 15V (voltage) & $\pm$ 10V  & $\pm$ 120V                        & $\pm$ 45V                & $\pm$ 100V     \\
-                                & $\pm$ 13.7V (current)    &       &                                   &                          &                \\
- Fastest pulse                  & 30$\mu$s      & 100 $\mu$s       & 20 $\mu$s                         & 60  $\mu$s               & 10 $\mu$s      \\
+ Compliance voltage             & $\pm$15~V (voltage) & $\pm$10~V  & $\pm$120~V                        & $\pm$45~V                & $\pm$100~V     \\
+                                & $\pm$13.7~V (current)    &       &                                   &                          &                \\
+ Fastest pulse                  & 30~$\mu$s      & 100~$\mu$s       & 20~$\mu$s                         & 60~$\mu$s               & 10~$\mu$s      \\
  Cost (USD)                     & \$194 (parts) & \$264 (parts)    & \$4131 (2 channels)               &  \$6211 (+ \$4438     & \$1708         \\
@@
 
-\textbf{Benchmarking.} To benchmark Stimjim, we varied the pulse frequency (from \Sexpr{round(1e6/max(paramMat$period),2)} to \Sexpr{round(1e6/min(paramMat$period),2)} Hz), duration (from \Sexpr{min(paramMat$dur0)} to \Sexpr{max(paramMat$dur0)} $\mu$s), and amplitude while simultaneously recording from both output channels on a National Instruments PCI-6110 card (12-bit ADC, sampling each channel at 2MHz).
-We set one channel to voltage mode and the other channel to current mode with a 9.75 k$\Omega$ resistor connected to the output.
+\textbf{Benchmarking.} To benchmark Stimjim, we generated a series of one-second biphasic pulse trains in which we varied the pulse frequency (from \Sexpr{round(1e6/max(paramMat$period),2)}~Hz to \Sexpr{round(1e6/min(paramMat$period),2)}~Hz), duration (from \Sexpr{min(paramMat$dur0)}~$\mu$s to \Sexpr{max(paramMat$dur0)}~$\mu$s), and amplitude.
+We simultaneously recording from both of Stimjim's output channels using a National Instruments PCI-6110 card (2~MHz sampling rate per channel, 4.9~mV resolution).
+One Stimjim channel was set to voltage mode and the other channel to current mode with a 9.75 k$\Omega$ resistor connected to the output.
 
@@
 \ref{fig:benchmarking}A-C).
-The standard deviation of interpulse intervals within a 1-second pulse train was always below one microsecond.
+The standard deviation of interpulse intervals within a one-second pulse train was always below one microsecond.
 Stimjim also produced pulse widths that were accurate and minimally variable (Fig \ref{fig:benchmarking}D-F).
 We then assessed pulse amplitudes across a range of settings to ensure negligible DC offsets and proper gains.
-From -10V to +10V (the range of our test equipment), Stimjim produced accurate voltage and current amplitudes, with maximal errors of less than 10mV and 3uA (Fig \ref{fig:benchmarking}G-I).
+From -10V to +10V (the range of our test equipment), Stimjim produced accurate voltage and current amplitudes, with maximal errors of less than 10~mV and 3~$\mu$A (Fig \ref{fig:benchmarking}G-I).
 
 <<benchmarkFigure, echo=FALSE,fig.height=8,fig.width=5,fig.cap="\\label{fig:benchmarking} \\textbf{Electrical benchmarks show Stimjim provides microsecond temporal precision and millivolt- and microampere-scale amplitude precision.}  (A) Example pulse trains with varying inter-pulse interval (IPI).
-(B) Median IPI in a 1-second pulse train measured on a high-speed data acquisition device, plotted against target IPI.
+(B) Median IPI in a one-second pulse train measured on a high-speed data acquisition device, plotted against target IPI.
 Red dots show current output, black dots show voltage output.
@@
 (C) Boxplot of IPI errors measured in a 1-second pulse train.
-Note that the due to the 2MHz sampling rate of our test equipment, the box representing the interquartile range often has zero height.
+Note that the due to the 2~MHz sampling rate of our test equipment, the box representing the interquartile range often has zero height.
 Whiskers are drawn to most extreme values, irrespective of interquartile range.
@@
 (G) Example pulses with varying amplitude.
-(H) Error in amplitude of voltage pulses (100 us, 1 kHz) vs target amplitude.
+(H) Error in amplitude of voltage pulses (100~$\\mu$s, 1~kHz) vs target amplitude.
 (I) same as (H) but for current pulses.
-(J) Example of a fast biphasic (30 us/phase) pulse.
-Red line is a 500uA current pulse (with a 9.75k resistive load), black line is a 5V pulse.
+(J) Example of a fast biphasic (30~$\\mu$s/phase) pulse.
+Red line is a 500~$\\mu$A current pulse (with a 9.75 k$\\Omega$ resistive load), black line is a 5~V pulse.
 (K) Example of a small-amplitude pulse.
-Red line is a 20uA current pulse (with a 9.75k resistive load), black line is a 0.2V pulse.", fig.align='center', out.width='1\\linewidth', fig.env="figure">>=
+Red line is a 20~$\\mu$A current pulse (with a 9.75 k$\\Omega$ resistive load), black line is a 0.2~V pulse.", fig.align='center', out.width='1\\linewidth', fig.env="figure">>=
 
@@
 padj=-6
-layout(matrix(1:12,4,3,byrow=TRUE), widths=c(0.8,1,1))
+layout(matrix(c(rep(1:9, each=2),rep(10,3), rep(11,3)),4,6,byrow=TRUE), widths=c(0.8,0.8,1.2,0.8,1,1))
 par(mar=c(4,1,2,1), mgp=c(2,0.3,0), cex.main=1.05, tck=-0.01,cex.axis=0.9)
@@
 mtext("B",side=2, padj=padj,las=2)
+legend('topleft', col=1:2, pch=1, legend=c("voltage","current"), bty='n')
 
@@
 mtext("E",side=2, padj=padj,las=2)
+legend('topleft', col=1:2, pch=1, legend=c("voltage","current"), bty='n')
 
@@
 
-par(mar=c(4,2,2,1), mgp=c(1,0.3,0))
+par(mar=c(4,3,2,1), mgp=c(2,0.3,0))
 dat = matrix(readBin(dataFiles[32], what='integer', n=1e7, size=2, endian = 'big'),ncol=2,byrow=TRUE)
 fp = min(which(dat[,1] > 100))
-ex4 = dat[fp + (-1e2:2e2),]*vpau
-matplot(1:nrow(ex4)/fs*1e6-50,ex4, main = 'Example fast pulse', type='l', lty=1, xlab='Time (us)', ylab='Amplitude (V)', yaxt='n')
+ex4 = dat[fp + (-60:160),]*vpau
+matplot(1:nrow(ex4)/fs*1e6-30, ex4, main = 'Example fast pulse', type='l', lty=1, xlab='Time (us)', ylab='Amplitude (V)', yaxt='n', bty='l')
 axis(2,las=2)
 mtext("J",side=2, padj=padj,las=2, line=1)
+legend('bottomleft', col=1:2, lty=1, legend=c("voltage","current"), bty='n')
 
@@
 ex5 = dat[fp + (-1e2:5e2),]*vpau
-matplot(1:nrow(ex5)/fs*1e6-50, ex5, main = 'Example small pulse', type='l', lty=1, xlab='Time (us)', ylab='Amplitude (V)', yaxt='n')
+matplot(1:nrow(ex5)/fs*1e6-50, ex5, main = 'Example small pulse', type='l', lty=1, xlab='Time (us)', ylab='Amplitude (V)', yaxt='n', bty='l')
 axis(2,las=2)
 mtext("K",side=2, padj=padj,las=2)
+legend('bottomleft', col=1:2, lty=1, legend=c("voltage","current"), bty='n')
 
@@
 Here, we trained two mice in a head-fixed variant of ICSS, in which animals could lick a sensor in order to obtain brain stimulation reward (BSR).
-We used a capacitive sensor attached to a small metal pole to detect licking, and every lick triggered a stimulus pulse train (0.5 seconds, initially 150 Hz and the minimal current at which animals would respond).
+We used a capacitive sensor attached to a small metal pole to detect licking, and every lick triggered a stimulus pulse train (0.5 seconds, initially 150~Hz and the minimal current at which animals would respond).
 To initially encourage licking, we placed a small amount of peanut butter on the metal sensor.
-After ini tial licking was reinforced by BSR, animals would continue licking long after the peanut butter was gone, including during the next session in which no peanut butter was offered.
+After initial licking was reinforced by BSR, animals would continue licking long after the peanut butter was gone, including during the next session in which no peanut butter was offered.
 After animals had learned the licking behavior (usually within their first hour session), we varied the stimulation frequency and amplitude and assessed how it affected licking behavior.
@@
 
-Notably, we observed clear differences between the two animals studied here.
+We observed clear differences between the two animals studied here.
 Mouse 1 shows a rather linear response to either increasing frequency or increasing amplitude, whereas mouse 2 had a more "digital" response akin to passing an activation threshold.
-Notably, maximal licking rates were comparable between the two animals.
+However, maximal licking rates were comparable between the two animals.
 Such differences are likely due to electrode placement, although they may also reflect intrinsically different personalities between the two animals.
@@
 <<icssAnalysis, echo=FALSE, fig.cap="\\label{fig:icss}\\textbf{Stimjim enables measuring frequency- and amplitude-dependent responses in an intracranial self-stimulation paradigm.} Mice decreased their licking rates when the frequency and amplitude of the rewarding stimulation decreased.
-For each animal, three frequency sweeps (highest to lowest frequency, 1 minute per frequency) were performed in a 45-minute session.
+For each animal, three frequency sweeps (highest to lowest frequency, one minute per frequency) were performed in a 45-minute session.
 Amplitude sweeps were performed in the same way.
-Each dot indicates a single 1-minute trial, and the solid black line shows the mean of all three trials at that frequency or amplitude.", cache=TRUE, fig.height=5, fig.width=5, fig.align='center'>>=
+Each dot indicates a single one-minute trial, and the solid black line shows the mean of all three trials at that frequency or amplitude.", cache=TRUE, fig.height=5, fig.width=5, fig.align='center'>>=
 
-icss.files = c("/home/x/Desktop/stimjim/benchmarking/2019-04-04_4309/20190404_115638.txt",
-               "/home/x/Desktop/stimjim/benchmarking/2019-04-04_4309/20190404_143910.txt",
-               "/home/x/Desktop/stimjim/benchmarking/2019-04-04_4476/20190404_130105.txt",
-               "/home/x/Desktop/stimjim/benchmarking/2019-04-04_4476/20190404_181300.txt")
+icss.files = c("/home/x/Dropbox (Technion Dropbox)/stimjim/benchmarking/2019-04-04_4309/20190404_115638.txt",
+               "/home/x/Dropbox (Technion Dropbox)/stimjim/benchmarking/2019-04-04_4309/20190404_143910.txt",
+               "/home/x/Dropbox (Technion Dropbox)/stimjim/benchmarking/2019-04-04_4476/20190404_130105.txt",
+               "/home/x/Dropbox (Technion Dropbox)/stimjim/benchmarking/2019-04-04_4476/20190404_181300.txt")
 
@@
 mtext("Mouse 2", 2, adj=0.3,  outer=TRUE, cex=0.75)
-mtext("Frequency Sweep", 3, adj=0.25,  outer=TRUE, cex=0.75)
-mtext("Amplitude Sweep", 3, adj=0.85,  outer=TRUE, cex=0.75)
+mtext("Frequency Sweep", 3, adj=0.25,  outer=TRUE, cex=1)
+mtext("Amplitude Sweep", 3, adj=0.85,  outer=TRUE, cex=1)
 
@@
 \hline
- Slew rate (voltage mode)                    & 4.1         & V/$\mu$s     \\
- Slew rate (current mode)                    & 0.8         & V/$\mu$s     \\
- Output voltage noise (voltage mode)$^a$     & 1           & mV rms    \\
-                                             & 5           & mVpp      \\
- Output voltage noise (current mode)$^a$     & 0.6         & mV rms     \\
-                                             & 3           & mVpp      \\
+ Slew rate (voltage mode)                    & 4.1         & V$\cdot\mu$s$^{-1}$     \\
+ Slew rate (current mode)                    & 0.8         & V$\cdot\mu$s$^{-1}$     \\
+ Output voltage noise (voltage mode)$^a$     & 1           & mV~rms    \\
+                                             & 5           & mV~p-p      \\
+ Output voltage noise (current mode)$^a$     & 0.6         & mV~rms     \\
+                                             & 3           & mV~p-p      \\
  Min. voltage increment$^b$                  & 0.45        & V         \\
@@
 }
-\caption{\label{tab:stimjimSpecs} \textbf{Stimjim electrical characteristics.} $^a$Measured on Tektronix TDS210 digital oscilloscope, full bandwidth (60MHz), 25ms segments..
+\caption{\label{tab:stimjimSpecs} \textbf{Stimjim electrical characteristics.} $^a$Measured on Tektronix TDS210 digital oscilloscope, full bandwidth (60MHz), calculated for 25~ms segments.
 Note that this is the noise only when the pulse is delivered; at all other times the output is directly connected to ground.
@@
 $^b$Calculated as output range divided by resolution of output DAC (16 bit).
-$^c$ value from from datasheet for Vishay DG509B (output switch).
+$^c$value from from datasheet for Vishay DG509B (output switch).
 }
@@
 \section*{Methods}
-\textbf{Electrode implantation}.
-Monopolar electrodes (Plastics1, MS303/2-AIU/SPC, coated stainless steel, 200um diameter) were implanted above the medial forebrain bundle according to the protocol in reference \cite{carlezon_intracranial_2007}.
+\textbf{Electrode implantation and ICSS}.
+Monopolar electrodes (Plastics1, MS303/2-AIU/SPC, coated stainless steel, 200~$\mu$m diameter) were implanted above the medial forebrain bundle according to the protocol in reference \cite{carlezon_intracranial_2007}.
 The ground was implanted in the contralateral cortex.
 Additionally, a 3D-printed headpost was affixed to the animal's skull by dental cement to enable head fixation.
-Typical resistance between connector pins after implantation was 20-30k$\Omega$.
+Typical resistance between connector pins after implantation was 20-30~k$\Omega$.
 All animal procedures were in accordance with guidelines established by the NIH on the care and use of animals in research and were confirmed by the Technion Institutional Animal Care and Use Committee (IL-012-01-18, valid until 10/4/2022).
 
+\textbf{Stimjim fabrication and benchmarking}.
+PCBs for Stimjim were ordered from SeeedStudio and components were ordered from Digikey.
+Components were manually soldered to the PCB using solder paste and a soldering iron.
+After soldering, the pulse control program was downloaded to the Teensy using the Arduino IDE and Teensyduino.
+From that point on, Stimjim's settings were controlled via serial communication over USB.
+We used a custom NI LabView program to set Stimjim's pulse parameters (frequency, amplitude, duration, etc.), initiate a one-second pulse train, and record both Stimjim channels using a National Instruments PCI-6110 card via a breakout box.
+
 \section*{Supplemental materials}
-In case of further modifications, the most up-to-date details regarding Stimjim will be available at https://bitbucket.org/natecermak/stimjim.
+In case of future modifications, the most up-to-date details regarding Stimjim will be available at https://bitbucket.org/natecermak/stimjim.
 The following are included as supplemental materials for the version of Stimjim documented here (v0.17).

### 2019-06-17 lib/stimjim/src/Stimjim.h
basic documentation to functions in stimjim library

@@ -13,7 +13,7 @@
 
 #define MICROAMPS_PER_DAC 0.1017            // 20V * (1/(3000 V/A)) / 2^16 = 0.1uA / DAC unit
 #define MICROAMPS_PER_ADC 0.85              //  (1/(100*(1+49.9k/1.8k) V/A)) * 20V / 2^13 = 0.85
-#define MILLIVOLTS_PER_DAC 0.4593            // 20V / 2^16 * gain of 1.505 ~= 0.44 mV / DAC unit
+#define MILLIVOLTS_PER_DAC 0.4593           // 20V / 2^16 * gain of 1.505 ~= 0.44 mV / DAC unit
 #define MILLIVOLTS_PER_ADC 2.44             // 20V / 2^13 = 2.44 mV / ADC unit
 
 #define OE0_1 6
@@ -39,21 +39,48 @@
 class StimJim {
 
   public:
+	// begin() should be run before using any other StimJim methods. 
+	// It initializes the SPI, creates appropriate SPISettings objects for DACs and ADCs,
+	// sets all necessary pins to outputs, sets ADC range to +-10V, and measures current offsets.
     void begin();
+	
+	// setOutputMode() sets the output mode for the selected channel (0/1).
+	// mode 0-3 are: VOLTAGE, CURRENT, HIGH-Z, GROUND.
     void setOutputMode(byte channel, byte mode);
+	
+	// Sets ADC range on both channels to +-2.5V, +-5V, or +-10V (use range=2.5, 5 
+	// or 10 to select, respectively).
     void setADCrange(float range);
+	
+	// Write a 16-bit signed int value to each channel. Outputs will be nearly 
+	// synchronously updated (sub-microsecond). 
     void writeToDacs(int16_t amp0, int16_t amp1);
+	
+	// Write a 16-bit signed int value to a single channel.     
     void writeToDac(byte channel, int16_t amp);
+	
+	// read the ADC on for a channel. line=0 means measure voltage at output,
+	// line=1 means read the value from the current sense instrument amplifier.
     int readADC(byte channel, byte line);
+	
+    // getCurrentOffsets() sweeps through a range of DAC values while measuring the
+	// current shunted through a 1k resistor (output remains grounded). This enables
+	// compensation for imperfect resistor matching and subsequent DC offset in the 
+	// Howland current pump.
     void getCurrentOffsets();
+
+    // getVoltageOffsets() sweeps through a range of DAC values while measuring the 
+	// output voltage. 
[…]

### 2019-06-30 benchmarking/paper/draft2.Rnw
sped up trigger response in stimjimPulser to 14us. updated photo. replaced many digitalWrites with digitalWriteFast in library. updates to analysis.

@@
 \usepackage[font={small}]{caption}
+
+%\addtolength{\textfloatsep}{-0.2in}
+%\addtolength{\dblfloatsep}{-0.2in}
 
@@
 
-\title{Stimjim is open source hardware for precise electrical stimulation}
+\title{Stimjim: open source hardware for precise electrical stimulation}
 \author{}
@@
 
-<<loadBenchmarkData, echo=FALSE, cache=TRUE>>=
+<<loadBenchmarkData, echo=FALSE, cache=TRUE,message=FALSE>>=
 library(parallel)
+library(readxl)
+library(gplots)
 
@@
 
-BOM=openxlsx::read.xlsx("../stimjim_BOM.xlsx")
-stimjimCost = ceiling(sum(BOM$Unit.cost*BOM$Quantity, na.rm=TRUE))
+BOM=readxl::read_excel("../stimjim_BOM.xlsx")
+stimjimCost = ceiling(sum(BOM[, 'Unit cost']*BOM$Quantity, na.rm=TRUE))
 
@@
 \section*{Abstract}
-Electrical stimulation is a simple and powerful tool to perturb and understand the function of neurons and neural circuits.
+Electrical stimulation is a simple and powerful tool to perturb and evoke neuronal activity in order to understand the function of neurons and neural circuits.
 Despite this, devices that can provide precise current or voltage stimulation are expensive and closed-source.
@@
 Electrical stimulation of neural tissue is an invaluable and ubiquitous research tool.
-Over the past 150 years, it has enabled an understanding of the function of various brain regions by directly causing neurons in those regions to fire \cite{hitzig_uber_1870,ferrier_functions_1876,penfield_somatic_1937}.
+Over the past 150 years, it has helped researchers understand the  function of various brain regions by directly inducing neurons in those regions to fire \cite{hitzig_uber_1870,ferrier_functions_1876,penfield_somatic_1937}.
 More recently, it has also found important clinical applications in neurological disorders including Parkinson's disease\cite{deuschl_randomized_2006} and depression\cite{mayberg_deep_2005}.
-However, to date, the hardware enabling precise control over current and voltage stimulation generally remains expensive and closed source.
+However, to date, the hardware for performing precise current- and voltage-based electrical stimulation generally remains expensive and closed source.
 
-In contrast, there has been a recent push within the scientific community to produce open labware - open source hardware and software replacements for a variety of common laboratory tasks\cite{pearce_open-source_2013,baden_open_2015}.
+In contrast, there has been a recent push within the scientific community to produce open labware - open source hardware and software replacements for a variety of common laboratory tasks\cite{pearce_open-source_2013,baden_open_2015;siegle_neural_2015}.
 Examples in life sciences include software and hardware for:
 \begin{itemize*}
-\item recording or stimulating neurons (e.g.
-Open Ephys\cite{siegle_open_2017,buccino_open_2018}, Cyclops\cite{newman_optogenetic_2015}, Neurorighter\cite{newman_closed-loop_2013}, Bonsai\cite{lopes_bonsai:_2015})
+\item recording or stimulating neurons (e.g. Open Ephys\cite{siegle_open_2017,buccino_open_2018}, Cyclops\cite{newman_optogenetic_2015}, Neurorighter\cite{newman_closed-loop_2013})
 \item amplifying DNA (e.g., OpenPCR\cite{noauthor_openpcr_nodate}) or preparing single cells for DNA sequencing \cite{booeshaghi_design_2019}
@@
 We now add Stimjim to this growing body of open hardware.
-Stimjim replaces commercial \textit{in vivo} neural stimulators at a fraction of the cost, without compromising on quality.
+Stimjim replaces commercial \textit{in vivo} neural stimulators at a fraction of the cost, with improved programmability.
 Furthermore, due to its entirely open design and software, Stimjim can be modified by end users to fit their specific needs.
-
-\section*{Results}
-\textbf{Design.} We developed Stimjim to be a precise, isolated stimulus generator.
-Stimjim is based on the Teensy 3.5 microcontroller, which utilizes a 32-bit Arm Cortex-M4 processor running at 120MHz.
-Each stimulating channel includes a current source based on an improved Howland current pump\cite{texas_instruments_-1515_2008}, and a voltage source (an op-amp), driven by a 16-bit digital-to-analog converter (DAC).
-The final output of each channel is selected by a 4-way switch, such that either channel can be configured as a current output, voltage output, grounded, or disconnected.
-To ensure the stimulator is properly connected (a common issue with experiments in freely moving animals) and to verify required stimulus current or voltage amplitudes, each channel also has a analog-to-digital converter (ADC) able to read either the output voltage or the output current (via a low-value sense resistor in series with the current output).
-Our circuit design was made in Kicad, an open-source printed circuit board (PCB) design program.
-Schematic, layout, bill of materials, and build instructions are included as supplemental materials and are also available in a git repository (https://bitbucket.org/natecermak/stimjim).
 
@@
 \begin{centering}
-\includegraphics[width=0.85\linewidth]{../../photo.png}
+\includegraphics[width=0.75\linewidth]{../../photo.png}
 \caption{\textbf{Stimjim is a compact, configurable, and precise stimulator.}}
@@
 
-Stimjim's design compares well against alternatives (Table 1).
-It is order-of-magnitude less expensive than most commercial alternatives.
+\section*{Results}
+\textbf{Design.} We developed Stimjim to be a precise, electrically isolated stimulus generator.
+Stimjim is based on the Teensy 3.5 microcontroller, which utilizes a 32-bit Arm Cortex-M4 processor running at 120MHz.
+Each stimulating channel includes a current source based on an improved Howland current pump\cite{texas_instruments_-1515_2008}, and a voltage source (an op-amp), driven by a 16-bit digital-to-analog converter (DAC).
+The final output of each channel is selected by a 4-way switch, such that either channel can be configured as a current output, voltage output, grounded, or disconnected.
+To ensure the stimulator is properly connected (a common issue with experiments in freely moving animals) and to verify required stimulus current or voltage amplitudes, each channel also has a analog-to-digital converter (ADC) able to read either the output voltage or the output current (via a low-value sense resistor in series with the current output).
+Our circuit board design was made using Kicad, an open-source printed circuit board (PCB) design program.
+Schematic, layout, bill of materials, and build instructions are included as supplemental materials and are also available in the Stimjim git repository (https://bitbucket.org/natecermak/stimjim).
+
+Stimjim's design compares favorably against alternatives (Table 1).
+It is an order of magnitude less expensive than most commercial alternatives.
 In comparison with commercial stimulators, its only drawback is that its compliance voltage is lower.
-This limits the load resistance that Stimjim can drive - for a given resistance $R$, Stimjim cannot output a current larger than $\frac{13.7~V}{R}$.
+This limits the load resistance that Stimjim can drive - for a given resistance $R$, Stimjim cannot output a current larger than $\frac{\mbox{13.7~V}}{R}$.
+Thus, Stimjim is not suitable for high-impedance electrodes such as pulled glass electrodes.
 Stimjim's cost is most similar to PulsePal 2, an open-source programmable pulse generator \cite{sanders_low-cost_2014}.
@@
 
-Stimjim's software is written in the Arduino language (C++) and is compatible with the Arduino programming environment.
-We provide an Arduino library consisting of low-level functions (writing registers in the DACs or ADCs, or setting the channel mode).
+Stimjim's software is written in C++ using the the Arduino development environment.
+We provide an Arduino-compatible Stimjim library permitting low-level device control (writing registers in the DACs or ADCs, or setting the stimulation control mode).
 Library functions enable users to create new programs to run on the Stimjim - for example, generation of custom waveform outputs stored on the onboard SD card.
-We also provide a default program designed to output pulse trains.
-Pulse trains are parameterized based on their output mode (current or voltage), frequency, duration, and the amplitude of each phase of the pulse itself.
+We also provide a default program that can generate user-defined pulse train sequences.
+Users set the parameters for pulse trains, including output mode (current or voltage), frequency, duration, and the amplitude of each phase of the pulse itself.
+Stimjim can store definitions for 100 pulse trains concurrently, and users can select and initiate particular pulse trains on the fly.
 
@@
                                 & $\pm$13.7~V (current)    &       &                                   &                          &                \\
- Fastest pulse                  & 30~$\mu$s      & 100~$\mu$s       & 20~$\mu$s                         & 60~$\mu$s               & 10~$\mu$s      \\
+ Fastest pulse                  & $\approx$30~$\mu$s      & 100~$\mu$s       & 20~$\mu$s                         & 60~$\mu$s               & 10~$\mu$s      \\
  Cost (USD)                     & \$194 (parts) & \$264 (parts)    & \$4131 (2 channels)               &  \$6211 (+ \$4438     & \$1708         \\
@@
 
-\textbf{Benchmarking.} To benchmark Stimjim, we generated a series of one-second biphasic pulse trains in which we varied the pulse frequency (from \Sexpr{round(1e6/max(paramMat$period),2)}~Hz to \Sexpr{round(1e6/min(paramMat$period),2)}~Hz), duration (from \Sexpr{min(paramMat$dur0)}~$\mu$s to \Sexpr{max(paramMat$dur0)}~$\mu$s), and amplitude.
-We simultaneously recording from both of Stimjim's output channels using a National Instruments PCI-6110 card (2~MHz sampling rate per channel, 4.9~mV resolution).
-One Stimjim channel was set to voltage mode and the other channel to current mode with a 9.75 k$\Omega$ resistor connected to the output.
-
-Stimjim proved exceedingly accurate, providing microsecond temporal resolution and millivolt- and microampere- amplitude resolution.
-Across the tested range of stimulation frequencies, Stimjim provided accurate and highly consistent inter-pulse intervals (IPIs; Fig.
-\ref{fig:benchmarking}A-C).
-The standard deviation of interpulse intervals within a one-second pulse train was always below one microsecond.
-Stimjim also produced pulse widths that were accurate and minimally variable (Fig \ref{fig:benchmarking}D-F).
-We then assessed pulse amplitudes across a range of settings to ensure negligible DC offsets and proper gains.
-From -10V to +10V (the range of our test equipment), Stimjim produced accurate voltage and current amplitudes, with maximal errors of less than 10~mV and 3~$\mu$A (Fig \ref{fig:benchmarking}G-I).
-
-<<benchmarkFigure, echo=FALSE,fig.height=8,fig.width=5,fig.cap="\\label{fig:benchmarking} \\textbf{Electrical benchmarks show Stimjim provides microsecond temporal precision and millivolt- and microampere-scale amplitude precision.}  (A) Example pulse trains with varying inter-pulse interval (IPI).
+<<benchmarkFigure, echo=FALSE,fig.height=8,fig.width=5,warning=FALSE,message=FALSE,fig.cap="\\label{fig:benchmarking} \\textbf{Electrical benchmarks show Stimjim provides microsecond temporal precision and millivolt- and microampere-scale amplitude precision.}  (A) Example pulse trains with varying inter-pulse interval (IPI).
 (B) Median IPI in a one-second pulse train measured on a high-speed data acquisition device, plotted against target IPI.
@@
   fp = min(which(dat[,1] > 100))
-  dat[fp + (-1e3:3e4),1]*vpau
+  dat[fp + (-1e3:2.9e4),1]*vpau
 })
@@
 
-boxplot(lapply(seq(1,60), function(i) res[[i]]$IPI*1e6 - paramMat[(i+1)/2,3]), axes=FALSE, range=0, lty=1, staplewex=0,
-  log='', main='', cex=0.75, xlab='Target IPI (ms)', ylab='IPI error (us)', col=1:2, border=1:2, ylim=c(-4,4))
+x = rep(paramMat[1:30,3]/1e3, each=2)
+y = sapply(seq(1,60), function(i) quantile(res[[i]]$IPI*1e6 - paramMat[(i+1)/2,3], c(0,0.5,1)))
+
+gplots::plotCI(x,y[2,], li=y[1,], ui=y[3,], yaxt='n', lty=1, gap=0, sfrac=0,
+  log='x', main='', cex=0.75, xlab='Target IPI (ms)', ylab='IPI error (us)', col=1:2, ylim=c(-4,4))
 axis(2,las=2)
-axis(1,las=2, at=seq(1.5,60,by=2), labels=signif(paramMat[1:30,3]/1e3, 3))
 mtext("C",side=2, padj=padj,las=2)
@@
 par(mar=c(4,3,2,1))
-plot(paramMat[31:59,7], sapply(res[seq(61,118,by=2)], function(x) x$PW.summary[3])*1e6/2,  yaxt='n',
-  log='xy', main='', xlab='Target PW (us)', ylab='Measured PW (us)', cex=0.75)
-points(paramMat[31:59,7], sapply(res[seq(62,118,by=2)], function(x) x$PW.summary[3])*1e6/2,
+plot(paramMat[31:59,7]/1e3, sapply(res[seq(61,118,by=2)], function(x) x$PW.summary[3])*1e3/2,  yaxt='n',
+  log='xy', main='', xlab='Target PW (ms)', ylab='Measured PW (us)', cex=0.75)
+points(paramMat[31:59,7]/1e3, sapply(res[seq(62,118,by=2)], function(x) x$PW.summary[3])*1e3/2,
   col='red', cex=0.75)
@@
 
-boxplot(lapply(seq(61,118), function(i) res[[i]]$PW*1e6/2 - paramMat[(i+1)/2,7]), axes=FALSE, range=0, lty=1, staplewex=0,
-  log='', main='', cex=0.75, xlab='Target PW (ms)', ylab='PW error (us)', col=1:2, border=1:2)
+x = rep(paramMat[31:60,7]/1e3, each=2)
+y = sapply(seq(61,120), function(i) quantile(res[[i]]$PW*1e6/2 - paramMat[(i+1)/2,7], c(0,0.5,1)))
+gplots::plotCI(x,y[1,], li=y[1,], ui=y[3,], yaxt='n', range=0, lty=1, gap=0, sfrac=0,
+  log='x', main='', cex=0.75, xlab='Target PW (ms)', ylab='PW error (us)', col=1:2)
 axis(2,las=2)
-axis(1,las=2, at=seq(1.5,60,by=2), labels=signif(paramMat[1:30,3]/1e3, 3))
 mtext("F",side=2, padj=padj,las=2)
@@
   fp = min(which(dat[,1] > 10))
-  dat[fp + (-200:700),1]*vpau
+  dat[fp + (-200:600),1]*vpau
 })
@@
 
-par(mar=c(4,3,2,1), mgp=c(2,0.3,0))
+par(mar=c(4,3,2,4), mgp=c(2,0.3,0))
 dat = matrix(readBin(dataFiles[32], what='integer', n=1e7, size=2, endian = 'big'),ncol=2,byrow=TRUE)
@@
 ex4 = dat[fp + (-60:160),]*vpau
-matplot(1:nrow(ex4)/fs*1e6-30, ex4, main = 'Example fast pulse', type='l', lty=1, xlab='Time (us)', ylab='Amplitude (V)', yaxt='n', bty='l')
+matplot(1:nrow(ex4)/fs*1e6-30, ex4, main = 'Example fast pulse', type='l', lty=1, xlab='Time (us)', ylab='Amplitude (V)', yaxt='n')
 axis(2,las=2)
@@
 legend('bottomleft', col=1:2, lty=1, legend=c("voltage","current"), bty='n')
+currentLabels = seq(-500,500,by=100)
+axis(4, las=2, at = currentLabels/1e6*9750, labels = currentLabels, col='red', col.axis='red'); 
+corners = par("usr"); par(xpd = TRUE);
+text(x = corners[2]+25.5, y = mean(corners[3:4]), 'Amplitude (uA)', srt = 270, col='red')
 
-par(mar=c(4,3,2,1), mgp=c(2,0.3,0))
+par(mar=c(4,3,2,4), mgp=c(2,0.3,0))
 dat = matrix(readBin(dataFiles[61], what='integer', n=1e7, size=2, endian = 'big'),ncol=2,byrow=TRUE)
@@
 ex5 = dat[fp + (-1e2:5e2),]*vpau
-matplot(1:nrow(ex5)/fs*1e6-50, ex5, main = 'Example small pulse', type='l', lty=1, xlab='Time (us)', ylab='Amplitude (V)', yaxt='n', bty='l')
+matplot(1:nrow(ex5)/fs*1e6-50, ex5, main = 'Example small pulse', type='l', lty=1, xlab='Time (us)', ylab='Amplitude (V)', yaxt='n')
 axis(2,las=2)
@@
 legend('bottomleft', col=1:2, lty=1, legend=c("voltage","current"), bty='n')
+currentLabels = seq(-20,20,by=5)
+axis(4, las=2, at = currentLabels/1e6*9750, labels = currentLabels, col='red', col.axis='red'); 
+corners = par("usr"); par(xpd = TRUE);
+text(x = corners[2]+55.5, y = mean(corners[3:4]), 'Amplitude (uA)', srt = 270, col='red')
 
 @
+
+\textbf{Benchmarking.} To benchmark Stimjim, we generated a series of one-second biphasic pulse trains in which we varied the pulse frequency (from \Sexpr{round(1e6/max(paramMat$period),2)}~Hz to \Sexpr{round(1e6/min(paramMat$period),2)}~Hz), duration (from \Sexpr{min(paramMat$dur0)}~$\mu$s to \Sexpr{max(paramMat$dur0)}~$\mu$s), and amplitude.
+We simultaneously recording from both of Stimjim's output channels using a National Instruments PCI-6110 card (2~MHz sampling rate per channel, 4.9~mV resolution).
+One Stimjim channel was set to voltage mode and the other channel to current mode with a 9.75 k$\Omega$ resistor connected to the output.
+
+Stimjim proved capable of providing microsecond temporal resolution and millivolt- and microampere- amplitude resolution.
+Across the tested range of stimulation frequencies, Stimjim generated accurate and highly consistent inter-pulse intervals (IPIs; Fig.
+\ref{fig:benchmarking}A-C).
+The standard deviation of inter-pulse intervals within a one-second pulse train was always below one microsecond.
+Stimjim also produced pulse widths that were accurate and minimally variable (Fig \ref{fig:benchmarking}D-F).
+For both inter-pulse interval and pulse width, the error in their durations increased as the duration itself increased (Fig.
+\ref{fig:benchmarking}C,F), but remained below 0.0004\% (IPI error) and 0.002\% (PW error).
+We then assessed pulse amplitudes across a range of settings to ensure negligible DC offsets and proper gains.
+From -10V to +10V (the range of our test equipment), Stimjim produced accurate voltage and current amplitudes, with maximal errors of less than 10~mV and 3~$\mu$A (Fig \ref{fig:benchmarking}G-I).
 
@@
 To initially encourage licking, we placed a small amount of peanut butter on the metal sensor.
-After initial licking was reinforced by BSR, animals would continue licking long after the peanut butter was gone, including during the next session in which no peanut butter was offered.
+After initial licking was reinforced by BSR, animals would continue licking long after the peanut butter was gone, including during the next session in which no peanut buwtter was offered.
 After animals had learned the licking behavior (usually within their first hour session), we varied the stimulation frequency and amplitude and assessed how it affected licking behavior.
@@
 
+\begin{table}
+\small{
+\begin{tabular}{|l|c|c|}
+\hline
+                                            & Value     & Units     \\
+\hline
+ Slew rate (voltage mode)                    & 4.1         & V$\cdot\mu$s$^{-1}$     \\
+ Slew rate (current mode)                    & 0.8         & V$\cdot\mu$s$^{-1}$     \\
+ Output voltage noise (voltage mode)$^a$     & 1           & mV~rms    \\
+                                             & 5           & mV~p-p      \\
+ Output voltage noise (current mode)$^a$     & 0.6         & mV~rms     \\
+                                             & 3           & mV~p-p      \\
+ Min. voltage increment$^b$                  & 0.45        & mV         \\
+ Min. current increment$^b$                  & 0.1         & $\mu$A        \\ 
+ Output impedance (voltage mode)$^c$         & 180         & $\Omega$ \\ 
+ Trigger latency                             & 14          & $\mu$s        \\
+ \hline
+\end{tabular}
+}
+\caption{\label{tab:stimjimSpecs} \textbf{Stimjim electrical characteristics.} $^a$Measured on Tektronix TDS210 digital oscilloscope, full bandwidth (60MHz), calculated for 25~ms segments.
+Note that this is the noise only when the pulse is delivered; at all other times the output is directly connected to ground.
+Voltage noise in current mode was measured with a 9.75 k$\Omega$ resistive load.
+$^b$Calculated as output range divided by resolution of output DAC (16 bit).
+$^c$value from from datasheet for Vishay DG509B (output switch).
+}
+\end{table}
+
+\section*{Conclusion}
+We have introduced Stimjim, an inexpensive yet precise open-source stimulator for neuroscience.
+At a cost of roughly \$200 USD for parts, Stimjim is order of magnitude less expensive than commercial, proprietary alternatives.
+It offers microsecond temporal control of current and voltage with millivolt/microampere precision.
+
+Stimjim's low cost opens up many potential applications, such as learning paradigms that involve direct electrical stimulation.
+Stimjim's open source nature makes it straightforward for researchers to customize the stimulation parameters and use Stimjim in closed loop experiments.
+Furthermore, researchers who were previously limited to training only one animal at a time due to possessing only a single stimulator could now train 10 animals simultaneously for comparable cost.
+
 <<icssAnalysis, echo=FALSE, fig.cap="\\label{fig:icss}\\textbf{Stimjim enables measuring frequency- and amplitude-dependent responses in an intracranial self-stimulation paradigm.} Mice decreased their licking rates when the frequency and amplitude of the rewarding stimulation decreased.
@@
 Amplitude sweeps were performed in the same way.
-Each dot indicates a single one-minute trial, and the solid black line shows the mean of all three trials at that frequency or amplitude.", cache=TRUE, fig.height=5, fig.width=5, fig.align='center'>>=
+Each dot indicates a single one-minute trial, and the solid black line shows the mean of all three trials at that frequency or amplitude.", cache=TRUE, fig.height=5, fig.width=5, fig.align='center', fig.env="figure">>=
 
@@
 
-\begin{table}
-\small{
-\begin{tabular}{|l|c|c|}
-\hline
-                                            & Value     & Units     \\
-\hline
- Slew rate (voltage mode)                    & 4.1         & V$\cdot\mu$s$^{-1}$     \\
- Slew rate (current mode)                    & 0.8         & V$\cdot\mu$s$^{-1}$     \\
- Output voltage noise (voltage mode)$^a$     & 1           & mV~rms    \\
-                                             & 5           & mV~p-p      \\
- Output voltage noise (current mode)$^a$     & 0.6         & mV~rms     \\
-                                             & 3           & mV~p-p      \\
- Min. voltage increment$^b$                  & 0.45        & V         \\
- Min. current increment$^b$                  & 0.1         & $\mu$A        \\ 
- Output impedance (voltage mode)$^c$         & 180         & $\Omega$ \\ 
- Trigger latency                             & 23          & $\mu$s        \\
- \hline
-\end{tabular}
-}
-\caption{\label{tab:stimjimSpecs} \textbf{Stimjim electrical characteristics.} $^a$Measured on Tektronix TDS210 digital oscilloscope, full bandwidth (60MHz), calculated for 25~ms segments.
-Note that this is the noise only when the pulse is delivered; at all other times the output is directly connected to ground.
-Voltage noise in current mode was measured with a 9.75 k$\Omega$ resistive load.
-$^b$Calculated as output range divided by resolution of output DAC (16 bit).
-$^c$value from from datasheet for Vishay DG509B (output switch).
-}
-\end{table}
-
-\section*{Conclusion}
-We have introduced Stimjim, an inexpensive yet precise open-source stimulator for neuroscience.
-At a cost of roughly \$200 USD for parts, Stimjim is order of magnitude less expensive than commercial, proprietary alternatives.
-It offers microsecond temporal control of current and voltage with millivolt/microampere precision.
-
-Stimjim's low cost opens up many potential applications, such as learning paradigms that involve direct electrical stimulation.
-Stimjim's open source nature makes it straightforward for researchers to customize the stimulation parameters and use Stimjim in closed loop experiments.
-Furthermore, researchers who were previously limited to training only one animal at a time due to possessing only a single stimulator could now train 10 animals simultaneously for comparable cost.
-
 \section*{Acknowledgments}
@@
 \section*{Methods}
-\textbf{Electrode implantation and ICSS}.
-Monopolar electrodes (Plastics1, MS303/2-AIU/SPC, coated stainless steel, 200~$\mu$m diameter) were implanted above the medial forebrain bundle according to the protocol in reference \cite{carlezon_intracranial_2007}.
-The ground was implanted in the contralateral cortex.
-Additionally, a 3D-printed headpost was affixed to the animal's skull by dental cement to enable head fixation.
-Typical resistance between connector pins after implantation was 20-30~k$\Omega$.
-All animal procedures were in accordance with guidelines established by the NIH on the care and use of animals in research and were confirmed by the Technion Institutional Animal Care and Use Committee (IL-012-01-18, valid until 10/4/2022).
 
@@
 We used a custom NI LabView program to set Stimjim's pulse parameters (frequency, amplitude, duration, etc.), initiate a one-second pulse train, and record both Stimjim channels using a National Instruments PCI-6110 card via a breakout box.
+
+\textbf{Electrode implantation and ICSS}.
+Monopolar electrodes (Plastics1, #MS303/2-AIU/SPC, coated stainless steel, 200~$\mu$m diameter) were implanted above the medial forebrain bundle according to the protocol in reference \cite{carlezon_intracranial_2007}.
+The ground was implanted in the contralateral cortex.
+Additionally, a 3D-printed headpost was affixed to the animal's skull by dental cement to enable head fixation.
+Typical resistance between connector pins after implantation was 20-30~k$\Omega$.
+All animal procedures were in accordance with guidelines established by the NIH on the care and use of animals in research and were confirmed by the Technion Institutional Animal Care and Use Committee (IL-012-01-18, valid until 10/4/2022).
+
+\textbf{Field stimulation experiments}. Platinum-iridium electrodes (Alpha Omega, #387-102S01-11, 250~$\mu$m diameter, Parylene C and Polyamide coated, 0.1~M$\Omega$) were placed.
 
@@
 
+<<sliceExpt, eval=FALSE>>=
+if (!require(readABF)){
+  install.packages("devtools")
+  devtools::install_github("Zabolekar/abfReader")
+}
+
+sliceDat = readABF::readABF("./slice2/27_06_19_0016.abf")$data[[1]][,1]
+
+i = which(diff(sliceDat)>55)
+i = i[diff(c(i,1e9)) > 1]
+
+plot(sliceDat[1:5e5], type='l')
+abline(v=i[1:20], col='red')
+
+m = sapply(i, function(j) sliceDat[j+(-50:500)])
+image(m, zlim=c(-100,50))
+
+matplot(1:nrow(m)/10-5, dendRites::offset(m,10), type='l', lty=1, col=1, xlab='time (ms)', ylab='Membrane potential (mV)', yaxt='n',bty='n')
+
+aps = sapply(i, function(j) {
+  x = sliceDat[j+50:500]
+  length(getLocalMaxima(x,threshold=0))
+})
+aps = colSums(matrix(aps,nrow=5))
+@
+
 \end{document}

### 2019-06-30 lib/stimjim/src/Stimjim.cpp

@@ -14,15 +14,15 @@ void StimJim::begin(){
   // ------------- Turn on LEDS ----------------------------------- //
   pinMode(LED0, OUTPUT);
   pinMode(LED1, OUTPUT);
-  digitalWrite(LED0, HIGH);
-  digitalWrite(LED1, HIGH);
+  digitalWriteFast(LED0, HIGH);
+  digitalWriteFast(LED1, HIGH);
 
 
   // ------------- Setup output pins ------------------------------------ //
   int pins[] = {NLDAC_0, NLDAC_1, CS0_0, CS1_0, CS0_1, CS1_1, OE0_0, OE1_0, OE0_1, OE1_1, LED0, LED1};
   for (int i = 0; i < 12; i++) {
     pinMode(pins[i], OUTPUT);
-    digitalWrite(pins[i], HIGH);
+    digitalWriteFast(pins[i], HIGH);
   }
 
 
@@ -32,8 +32,8 @@ void StimJim::begin(){
   getCurrentOffsets();
   getVoltageOffsets();
   delay(1000);
-  digitalWrite(LED0, LOW);
-  digitalWrite(LED1, LOW);
+  digitalWriteFast(LED0, LOW);
+  digitalWriteFast(LED1, LOW);
 }
 
 void StimJim::setOutputMode(byte channel, byte mode) {

### 2019-07-01 benchmarking/paper/draft2.Rnw
added opa197 datasheet, updated analysis

@@
 \item amplifying DNA (e.g., OpenPCR\cite{noauthor_openpcr_nodate}) or preparing single cells for DNA sequencing \cite{booeshaghi_design_2019}
-\item imaging microscopic samples (e.g., FlyPi\cite{chagas_100_2017})
+\item imaging microscopic samples (e.g., FlyPi\cite{chagas_100_2017}, Octopi\cite{li_octopi:_2019})
 \item quantifying chemical reactions or microbial growth in 96-well plates\cite{szymula_open-source_2019}
@@
                                 & $\pm$13.7~V (current)    &       &                                   &                          &                \\
- Fastest pulse                  & $\approx$30~$\mu$s      & 100~$\mu$s       & 20~$\mu$s                         & 60~$\mu$s               & 10~$\mu$s      \\
+ Fastest pulse                  & $\sim$30~$\mu$s      & 100~$\mu$s       & 20~$\mu$s                         & 60~$\mu$s               & 10~$\mu$s      \\
  Cost (USD)                     & \$194 (parts) & \$264 (parts)    & \$4131 (2 channels)               &  \$6211 (+ \$4438     & \$1708         \\
@@
 Solid line shows equality.
-(C) Boxplot of IPI errors measured in a 1-second pulse train.
-Note that the due to the 2~MHz sampling rate of our test equipment, the box representing the interquartile range often has zero height.
-Whiskers are drawn to most extreme values, irrespective of interquartile range.
+(C) Plot of IPI errors measured in a 1-second pulse train.
+Points show median error, error bars indicate worst-case errors.
 (D) Example pulse trains with varying pulse width (PW).
@@
 Solid line shows equality.
-(F) Boxplots of pulse width errors over 100 pulses.
+(F) Plots of pulse width errors over 100 pulses.
 Conventions are the same as for (C).
@@
 par(mar=c(4,3,2,1))
-plot(paramMat[1:30,3]/1e3, sapply(res[seq(1,60,by=2)], function(x) x$IPI.summary[3])*1e3, yaxt='n',
+plot(paramMat[1:30,3]/1e3, sapply(res[seq(1,60,by=2)], function(x) x$IPI.summary[3])*1e3, xaxt='n',yaxt='n',
   log='xy', main='', cex=0.75, xlab='Target IPI (ms)', ylab='Measured IPI (ms)')
 axis(2,las=2)
+axis(1, las=2)
 points(paramMat[1:30,3]/1e3, sapply(res[seq(2,60,by=2)], function(x) x$IPI.summary[3])*1e3, 
@@
 
-gplots::plotCI(x,y[2,], li=y[1,], ui=y[3,], yaxt='n', lty=1, gap=0, sfrac=0,
+gplots::plotCI(x,y[2,], li=y[1,], ui=y[3,], lty=1, gap=0, sfrac=0,  xaxt='n',yaxt='n',
   log='x', main='', cex=0.75, xlab='Target IPI (ms)', ylab='IPI error (us)', col=1:2, ylim=c(-4,4))
+axis(1,las=2)
 axis(2,las=2)
@@
 par(mar=c(4,3,2,1))
-plot(paramMat[31:59,7]/1e3, sapply(res[seq(61,118,by=2)], function(x) x$PW.summary[3])*1e3/2,  yaxt='n',
+plot(paramMat[31:59,7]/1e3, sapply(res[seq(61,118,by=2)], function(x) x$PW.summary[3])*1e3/2, xaxt='n',yaxt='n',
   log='xy', main='', xlab='Target PW (ms)', ylab='Measured PW (us)', cex=0.75)
@@
 abline(a=0,b=1)
+axis(1,las=2)
 axis(2,las=2)
@@
 y = sapply(seq(61,120), function(i) quantile(res[[i]]$PW*1e6/2 - paramMat[(i+1)/2,7], c(0,0.5,1)))
-gplots::plotCI(x,y[1,], li=y[1,], ui=y[3,], yaxt='n', range=0, lty=1, gap=0, sfrac=0,
+gplots::plotCI(x,y[1,], li=y[1,], ui=y[3,], lty=1, gap=0, sfrac=0,  xaxt='n',yaxt='n',
   log='x', main='', cex=0.75, xlab='Target PW (ms)', ylab='PW error (us)', col=1:2)
+axis(1,las=2)
 axis(2,las=2)
@@
 \textbf{Electrode implantation and ICSS}.
-Monopolar electrodes (Plastics1, #MS303/2-AIU/SPC, coated stainless steel, 200~$\mu$m diameter) were implanted above the medial forebrain bundle according to the protocol in reference \cite{carlezon_intracranial_2007}.
+Monopolar electrodes (Plastics1, \#MS303/2-AIU/SPC, coated stainless steel, 200~$\mu$m diameter) were implanted above the medial forebrain bundle according to the protocol in reference \cite{carlezon_intracranial_2007}.
 The ground was implanted in the contralateral cortex.
@@
 
-\textbf{Field stimulation experiments}. Platinum-iridium electrodes (Alpha Omega, #387-102S01-11, 250~$\mu$m diameter, Parylene C and Polyamide coated, 0.1~M$\Omega$) were placed.
+\textbf{Field stimulation experiments}. Platinum-iridium electrodes (Alpha Omega, \#387-102S01-11, 250~$\mu$m diameter, Parylene C and Polyamide coated, 0.1~M$\Omega$) were placed.
 
@@
 
-<<sliceExpt, eval=FALSE>>=
+<<sliceExpt, eval=TRUE,echo=FALSE,message=FALSE>>=
 if (!require(readABF)){
@@
 
-sliceDat = readABF::readABF("./slice2/27_06_19_0016.abf")$data[[1]][,1]
+sliceDat = readABF::readABF(paste(projectDirectory,"slice2/27_06_19_0016.abf",sep=''))$data[[1]][,1]
 
@@
 
-plot(sliceDat[1:5e5], type='l')
-abline(v=i[1:20], col='red')
+#plot(sliceDat[1:5e5], type='l')
+#abline(v=i[1:20], col='red')
 
 m = sapply(i, function(j) sliceDat[j+(-50:500)])
-image(m, zlim=c(-100,50))
+image(x=1:551/10, z=m, zlim=c(-100,50),xlab='time (ms)')
 
-matplot(1:nrow(m)/10-5, dendRites::offset(m,10), type='l', lty=1, col=1, xlab='time (ms)', ylab='Membrane potential (mV)', yaxt='n',bty='n')
+#matplot(1:nrow(m)/10-5, dendRites::offset(m,10), type='l', lty=1, col=1, xlab='time (ms)', ylab='Membrane potential (mV)', yaxt='n',bty='n')
 
-aps = sapply(i, function(j) {
-  x = sliceDat[j+50:500]
-  length(getLocalMaxima(x,threshold=0))
-})
-aps = colSums(matrix(aps,nrow=5))
+#aps = sapply(i, function(j) {
+#  x = sliceDat[j+50:500]
+#  length(dendRites::getLocalMaxima(x,threshold=0))
+#})
+#aps = colSums(matrix(aps,nrow=5))
 @

### 2019-07-02 lib/stimjim/src/Stimjim.cpp
stimjim class now properly measures ADC offset and accounts for it. also enforced strict camelCase in class methods/fields.

@@ -6,19 +6,11 @@ void StimJim::begin(){
   SPI.begin();
 
   //AD5752 DAC runs at max 30MHz, SPI mode 1 or 2
-  SpiSettingsDAC = SPISettings(10000000, MSBFIRST, SPI_MODE1); 
+  SpiSettingsDac = SPISettings(10000000, MSBFIRST, SPI_MODE1); 
   //AD7321 settings - clock starts high, data latch on falling edge
-  SpiSettingsADC = SPISettings(5000000, MSBFIRST, SPI_MODE2);  
+  SpiSettingsAdc = SPISettings(5000000, MSBFIRST, SPI_MODE2);  
 
-
-  // ------------- Turn on LEDS ----------------------------------- //
-  pinMode(LED0, OUTPUT);
-  pinMode(LED1, OUTPUT);
-  digitalWriteFast(LED0, HIGH);
-  digitalWriteFast(LED1, HIGH);
-
-
-  // ------------- Setup output pins ------------------------------------ //
+  // ------------- Setup output pins, turn on LEDs ------------------------------------ //
   int pins[] = {NLDAC_0, NLDAC_1, CS0_0, CS1_0, CS0_1, CS1_1, OE0_0, OE1_0, OE0_1, OE1_1, LED0, LED1};
   for (int i = 0; i < 12; i++) {
     pinMode(pins[i], OUTPUT);
@@ -26,8 +18,11 @@ void StimJim::begin(){
   }
 
 
-  setADCrange(10);
-  setupDACs();
+  setupDacs();
+
+  getAdcOffsets();
+
+
 
   getCurrentOffsets();
   getVoltageOffsets();
@@ -43,9 +38,8 @@ void StimJim::setOutputMode(byte channel, byte mode) {
 }
 
 
-
-void StimJim::setADCrange(float range) {
-  SPI.beginTransaction(SpiSettingsADC);
+void StimJim::setAdcRange(float range) {
+  SPI.beginTransaction(SpiSettingsAdc);
   for (int i = 0; i < 2; i++) {
     int cs = (i == 0) ? CS1_0 : CS1_1; //CS1_x is ADC for channel x
     digitalWrite(cs, LOW); //[15] = write, [13] = range register
@@ -61,8 +55,8 @@ void StimJim::setADCrange(float range) {
   SPI.endTransaction();
 }
 
-void StimJim::setupDACs() {
-  SPI.beginTransaction(SpiSettingsDAC);
+void StimJim::setupDacs() {
[…]

### 2019-07-02 lib/stimjim/src/Stimjim.h

@@ -12,7 +12,7 @@
 #include <SPI.h>
 
 #define MICROAMPS_PER_DAC 0.1017            // 20V * (1/(3000 V/A)) / 2^16 = 0.1uA / DAC unit
-#define MICROAMPS_PER_ADC 0.85              //  (1/(100*(1+49.9k/1.8k) V/A)) * 20V / 2^13 = 0.85
+#define MICROAMPS_PER_ADC 0.85              // (1/(100*(1+49.9k/1.8k) V/A)) * 20V / 2^13 = 0.85
 #define MILLIVOLTS_PER_DAC 0.4593           // 20V / 2^16 * gain of 1.505 ~= 0.44 mV / DAC unit
 #define MILLIVOLTS_PER_ADC 2.44             // 20V / 2^13 = 2.44 mV / ADC unit
 
@@ -50,7 +50,7 @@ class StimJim {
 	
 	// Sets ADC range on both channels to +-2.5V, +-5V, or +-10V (use range=2.5, 5 
 	// or 10 to select, respectively).
-    void setADCrange(float range);
+    void setAdcRange(float range);
 	
 	// Write a 16-bit signed int value to each channel. Outputs will be nearly 
 	// synchronously updated (sub-microsecond). 
@@ -61,7 +61,12 @@ class StimJim {
 	
 	// read the ADC on for a channel. line=0 means measure voltage at output,
 	// line=1 means read the value from the current sense instrument amplifier.
-    int readADC(byte channel, byte line);
+    int readAdc(byte channel, byte line);
+	
+	// get ADC readings from ground (OUT_0 or OUT_1), when OUT_0 or OUT_1 is grounded. 
+	// this accounts for intrinsic "bipolar zero error" in AD7321 ADC.
+	void getAdcOffsets();
+	
 	
     // getCurrentOffsets() sweeps through a range of DAC values while measuring the
 	// current shunted through a 1k resistor (output remains grounded). This enables
@@ -74,14 +79,16 @@ class StimJim {
 	// Caution!! Anything connected to the output will see a voltage ramp!
 	// This enables compensation for any DC offset in the DAC or amplifier.
     void getVoltageOffsets();
+	float adcOffset25[2];   // read voltage on both channels at 2.5V range
+	float adcOffset10[2];    // read voltage on both channels at 10V range
 
     int currentOffsets[2];  // These store the compensation for various DC offsets from 
[…]

### 2019-07-02 benchmarking/makeTestParameters.R
updated test protocol, increased intervaltimer priority to 24

@@ -5,6 +5,7 @@
 
 projectDirectory = 'C:/Users/Jackie/Desktop/cermak/stimjim/'
 projectDirectory = '/home/x/Desktop/stimjim/'
+projectDirectory = 'D:/Dropbox (Technion Dropbox)/stimjim/'
 
 duration = 1e6
 paramMat = NULL
@@ -13,7 +14,7 @@ mode1 = 1
 amp = c(10,1)
 periods = round(exp(seq(log(250), log(5e5), length.out=30)))
 pulseDurations = round(exp(seq(log(25),log(4000), length.out=30)))
-ampScalings = seq(20,1000,length.out=30)
+ampScalings = seq(5,1000,length.out=30)
 
 ampScaling=500; pulseDuration=100
 for (period in periods){

### 2019-07-03 lib/stimjim/src/Stimjim.cpp
removed voltage sweep on boot. added way to check adc offsets and re-estimate them.

@@ -25,7 +25,7 @@ void StimJim::begin(){
 
 
   getCurrentOffsets();
-  getVoltageOffsets();
+  //getVoltageOffsets();
   delay(1000);
   digitalWriteFast(LED0, LOW);
   digitalWriteFast(LED1, LOW);

### 2019-07-04 lib/stimjim/src/Stimjim.cpp
fixed bug in startup that put ADCs in undefined state as to which channel they were reading

@@ -20,13 +20,14 @@ void StimJim::begin(){
 
   setupDacs();
 
+  setAdcLine(0,0);
+  setAdcLine(1,0);
+  
   getAdcOffsets();
-
-
-
   getCurrentOffsets();
   //getVoltageOffsets();
-  delay(1000);
+
+  delay(300);
   digitalWriteFast(LED0, LOW);
   digitalWriteFast(LED1, LOW);
 }
@@ -38,22 +39,6 @@ void StimJim::setOutputMode(byte channel, byte mode) {
 }
 
 
-void StimJim::setAdcRange(float range) {
-  SPI.beginTransaction(SpiSettingsAdc);
-  for (int i = 0; i < 2; i++) {
-    int cs = (i == 0) ? CS1_0 : CS1_1; //CS1_x is ADC for channel x
-    digitalWrite(cs, LOW); //[15] = write, [13] = range register
-    if (range == 2.5) {
-      SPI.transfer16(0b1011000100000000); // write range register +-2.5V on both channels
-    } else if (range == 5) {
-      SPI.transfer16(0b1010100010000000); // write range register +-5V on both channels)
-    } else {
-      SPI.transfer16(32768 + 8192);       // write range register, +-10V on both channels
-    }
-    digitalWrite(cs, HIGH);
-  }
-  SPI.endTransaction();
-}
 
 void StimJim::setupDacs() {
   SPI.beginTransaction(SpiSettingsDac);
@@ -109,6 +94,34 @@ void StimJim::writeToDac(byte channel, int16_t amp) {
   digitalWrite((channel) ? NLDAC_1 : NLDAC_0, HIGH);
 }
 
+
+
+void StimJim::setAdcRange(float range) {
+  SPI.beginTransaction(SpiSettingsAdc);
+  for (int i = 0; i < 2; i++) {
+    int cs = (i == 0) ? CS1_0 : CS1_1; //CS1_x is ADC for channel x
+    digitalWrite(cs, LOW); //[15] = write, [13] = range register
+    if (range == 2.5) {
+      SPI.transfer16(0b1011000100000000); // write range register +-2.5V on both channels
+    } else if (range == 5) {
+      SPI.transfer16(0b1010100010000000); // write range register +-5V on both channels)
+    } else {
+      SPI.transfer16(32768 + 8192);       // write range register, +-10V on both channels
+    }
+    digitalWrite(cs, HIGH);
[…]

### 2019-07-04 lib/stimjim/src/Stimjim.h

@@ -48,16 +48,18 @@ class StimJim {
 	// mode 0-3 are: VOLTAGE, CURRENT, HIGH-Z, GROUND.
     void setOutputMode(byte channel, byte mode);
 	
-	// Sets ADC range on both channels to +-2.5V, +-5V, or +-10V (use range=2.5, 5 
-	// or 10 to select, respectively).
-    void setAdcRange(float range);
-	
 	// Write a 16-bit signed int value to each channel. Outputs will be nearly 
 	// synchronously updated (sub-microsecond). 
     void writeToDacs(int16_t amp0, int16_t amp1);
 	
 	// Write a 16-bit signed int value to a single channel.     
     void writeToDac(byte channel, int16_t amp);
+
+	// Sets ADC range on both channels to +-2.5V, +-5V, or +-10V (use range=2.5, 5 
+	// or 10 to select, respectively).
+    void setAdcRange(float range);
+	
+	void setAdcLine(byte channel, byte line);
 	
 	// read the ADC on for a channel. line=0 means measure voltage at output,
 	// line=1 means read the value from the current sense instrument amplifier.
